---
title: "2-Preprocessing_Metabolomics"
author: "Cathy Magnée"
date: "2024-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prepare metabolomics data

This R Markdown document is a translation of the Rscript 'metabolomic.R', from Saezlabs Github page Factos_COSMOS, specifically the no_MOFA branch. This page can be found here <https://github.com/saezlab/Factor_COSMOS/tree/no_MOFA/scripts>


### Loading libraries

```{r}
library(readr)
library(reshape2)
library(pheatmap)
library(vsn)


# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("metaboliteIDmapping")

library(metaboliteIDmapping)

```

### Loading data

```{r}

# Metabolomic Data From Metabolon - data averaged from triplicate experiments (from https://wiki.nci.nih.gov/display/NCIDTPdata/Molecular+Target+Data)
# metabs <- as.data.frame(read_csv("data/metabolomic/WEB_DATA_METABOLON.TXT"))

# Metabolomic Data From Metabolon - individual data from each of the triplicate experiments (from https://wiki.nci.nih.gov/display/NCIDTPdata/Molecular+Target+Data)

# metabs <- as.data.frame(read_csv("data/metabolomic/WEB_DATA_METABOLON.TXT")) # Original file

metabs <- as.data.frame(read_csv("data/metabolomic/WEB_DATA_METABOLON_ALL.TXT")) # Added _ALL



```

Hopefully, one day an extra line of code, it being our data, will be added to this \<3

### Filter step I

First filtering step is to remove unclear metabolite measurements:

```{r}

# Remove unclear metabolite measurements
metabs <- metabs[!grepl("Isobar",metabs$TITLE),]
metabs <- metabs[!grepl("^X[-]",metabs$TITLE),]
metabs <- metabs[!grepl("^Possible",metabs$TITLE),]
```



```{r}

# metabs_df <- dcast(metabs, formula = TITLE~cellname, value.var = "VALUE",fun.aggregate = mean)
metabs_df <- dcast(metabs, formula = TITLE~cellname, value.var = "Value",fun.aggregate = mean) # For when using ALL metabolons
row.names(metabs_df) <- metabs_df$TITLE
metabs_df <- metabs_df[,-1]

```

## Plotting the distribution of metabolite values
Histogram:
```{r pressure, echo=FALSE}
# Inspect distribution of values
hist(as.numeric(unlist(log2(metabs_df))), breaks = 1000) #Things looks like they are getting funky around 5

# The above is Saezlab's commenting. IMHO, when looking at the graph below, things are mostly "funky" between -5 and 0. But maybe I do not have the required knowledge and experience about metabolomics data to understand what is funky about >5

```
Heatmap:
```{r}
pheatmap(log2(metabs_df), show_colnames = F, show_rownames = F) 

```
It is ugly and I do need help interpreting this tbh




### Removing outliers

```{r}

# Remove extreme values (outliers)
metabs_df[abs(log2(metabs_df)) > 5] <- NA 

```

### NO idea what they do here
```{r}
fit <- vsnMatrix(as.matrix(metabs_df))
meanSdPlot(fit)
metabs_df <- as.data.frame(vsn::predict(fit,as.matrix(metabs_df)))
```
Heatmap again:

```{r}
pheatmap(metabs_df[complete.cases(metabs_df),], show_colnames = T, show_rownames = F, cluster_rows = F) 

```
### Reformatting ?

```{r}
to_write <- metabs_df
to_write$metabolite <- row.names(to_write)
to_write <- to_write[,c(59,1:58)]
to_write$metabolite <- gsub(",","",to_write$metabolite)
```

### Saving
```{r}
write_csv(to_write, file = "data/metabolomic/metabolomic_clean_vsn.csv")
```

### ???
```{r}
samples <- names(metabs_df)
metabs <- row.names(metabs_df)

SDs <- apply(metabs_df,1,function(x){sd(x,na.rm = T)})
means <- rowMeans(metabs_df, na.rm = T)

metabs_df_scaled <- (metabs_df - means) / SDs
pheatmap(metabs_df_scaled[complete.cases(metabs_df_scaled),], show_colnames = T, show_rownames = F, cluster_rows = F) 

hist(as.numeric(unlist(metabs_df)), breaks = 1000) 
```

