---
title: "4-Run_CosmosR"
author: "Cathy Magnée, Saezlab"
date: "2024-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Run COSMOS
This document is the annotated version of the run_cosmos.R script that can be found on Saezlab's Factor_COSMOS github main branch, in the subfolder scripts, here: <https://github.com/saezlab/Factor_COSMOS/blob/main/scripts/run_cosmos.R>

### Installing packages, loading libraries, loading data

If you do not have them yet, please properly install CARNIVAL and COSMOS from Github:

```{r}
# remotes::install_github("saezlab/cosmosR")
# install.packages("devtools")
# devtools::install_github("saezlab/CARNIVAL")
```

Libraries:
```{r}
library(cosmosR)
library(readr)
library(dplyr)
library(CARNIVAL)
```
Loading data:
Select which cell line (for us, which patient??) you wish to analyze
```{r}
source("Factor_COSMOS/scripts/decoupleRnival.R")

data("meta_network")

meta_network <- meta_network_cleanup(meta_network)

load("Factor_COSMOS/data/cosmos/cosmos_inputs.RData")

```

### Selecting cell line to analyze
```{r}
names(cosmos_inputs)

cell_line <- "EKVX" # Select cell line you want to analyze with cosmos HERE


sig_input <- cosmos_inputs[[cell_line]]$TF_scores
metab_input <- cosmos_inputs[[cell_line]]$metabolomic
RNA_input <- cosmos_inputs[[cell_line]]$RNA
```

### Parameters

Choose which compartment to assign to the metabolic measurements:
Possible values are "r", "c", "e", "x", "m", "l", "n" and "g". More info here <http://bigg.ucsd.edu/compartments> 
"r", endoplasmic reticulum
"c", cytosol
"e", extracellular space
"x", peroxisome / glyoxysome
"m", mitochondria
"l", lysosome
"n", nucleus
"g", golgi apparatus

```{r}
#Choose which compartment to assign to the metabolic measurements
metab_input <- prepare_metab_inputs(metab_input, c("c","m"))

##Filter significant inputs
sig_input <- sig_input[abs(sig_input) > 2]
metab_input <- metab_input[abs(metab_input) > 2]

```


## IBM Cplex Options
```{r}
#In order to adapt options to users specification we can load them into a variable 
#that will then be passed to preprocess_COSMOS_signaling_to_metabolism CARNIVAL_options parameter
my_options <- default_CARNIVAL_options(solver = "cplex")

#Here the user should provide a path to its CPLEX executable (only cplex at the moment, other solvers will be documented soon !)
# my_options$solverPath <- "~/Documents/cplex" #or cbc solver executable
my_options$solverPath <- "C:/Program Files/IBM/ILOG/CPLEX_Studio2211/cplex/bin/x64_win64"
# my_options$solverPath <- "cbc/cbc-osx/cbc" #or cbc solver executable
my_options$solver <- "cplex" #or cbc
# my_options$solver <- "cbc"
my_options$timelimit <- 3600/10
my_options$mipGAP <- 0.05
my_options$threads <- 6
```
Note: You need IBM Cplex for this, your location may be different from mine. 


## Preprocessing signalling to metab

Use this only if preprocess_COSMOS_signaling_to_metabolism does not work:
```{r}
# source("Factor_COSMOS/scripts/decoupleRnival.R")
```


```{r}
metab_input <- cosmosR:::filter_input_nodes_not_in_pkn(metab_input, meta_network)
sig_input <- cosmosR:::filter_input_nodes_not_in_pkn(sig_input, meta_network)

test_for <- preprocess_COSMOS_signaling_to_metabolism(meta_network = meta_network[,c(1,3,2)],
                                                      signaling_data = sig_input,
                                                      metabolic_data = metab_input,
                                                      diff_expression_data = RNA_input,
                                                      maximum_network_depth = 4,
                                                      remove_unexpressed_nodes = T,
                                                      filter_tf_gene_interaction_by_optimization = T,
                                                      CARNIVAL_options = my_options)
```

