---
title: "4-Run_CosmosR"
author: "Cathy Magnée, Saezlab"
date: "2024-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Run COSMOS
This document is the annotated version of the run_cosmos.R script that can be found on Saezlab's Factor_COSMOS github main branch, in the subfolder scripts, here: <https://github.com/saezlab/Factor_COSMOS/blob/main/scripts/run_cosmos.R>

### Installing packages, loading libraries, loading data

If you do not have them yet, please properly install CARNIVAL and COSMOS from Github:

```{r}
# remotes::install_github("saezlab/cosmosR")
# install.packages("devtools")
# devtools::install_github("saezlab/CARNIVAL")
```

Libraries:
```{r}
library(cosmosR)
library(readr)
library(dplyr)
```
Loading data:
Select which cell line (for us, which patient??) you wish to analyze
```{r}
data("meta_network")

meta_network <- meta_network_cleanup(meta_network)

load("Factor_COSMOS/data/cosmos/cosmos_inputs.RData")

```

### Selecting cell line to analyze
```{r}
names(cosmos_inputs)

cell_line <- "786-0" # Select cell line you want to analyze with cosmos HERE


sig_input <- cosmos_inputs[[cell_line]]$TF_scores
metab_input <- cosmos_inputs[[cell_line]]$metabolomic
RNA_input <- cosmos_inputs[[cell_line]]$RNA
```

### Parameters

Choose which compartment to assign to the metabolic measurements:
Possible values are "r", "c", "e", "x", "m", "l", "n" and "g"
"r", 
"c", 
"e",
"x", 
"m", 
"l", 
"n",
"g",

```{r}
#Choose which compartment to assign to the metabolic measurments
metab_input <- prepare_metab_inputs(metab_input, c("c","m"))

##Filter significant inputs
sig_input <- sig_input[abs(sig_input) > 2]
metab_input <- metab_input[abs(metab_input) > 2]

```

