---
title: "2 - Preprocessing gene expression data"
author: "Cathy Magnée, Saezlab"
date: "2024-06-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preprocessing T cell transcriptomics data

This document describes the preprocessing of transcriptomic and metabolomic T cell data. It is based off the preprocessing documents in the FACTOR_Cosmos No_MOFA branch by Saezlab on github. This document is precedes the 2 - Preprocessing omics data.Rmd document. 

This document is annotated to explain the steps and the choices that were made durnig this project. Other options will be highlighted.

### Loading libraries
Installing packages (if you have not yet)

```{r}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("vsn")
# BiocManager::install("biomaRt")

```

Loading libraries:
```{r}
library(readxl)
library(readr)
library(pheatmap)
library(tibble)
```

Loading data
```{r}

# Defining paths
############ AMC ############  
# day0_RNA_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Transcriptomic\\day0_RNA_log2_FPKM.csv"
# 
# day2_RNA_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Transcriptomic\\day2_RNA_log2_FPKM.csv"

############ Laptop ############ 
day0_RNA_path <- "../ourdata/transcriptomic/day0_RNA_log2_FPKM.csv"
day2_RNA_path <- "../ourdata/transcriptomic/day2_RNA_log2_FPKM.csv"

# day0_RNA_path <- "../ourdata/transcriptomic/day0_RNA_raw.csv"
# day2_RNA_path <- "../ourdata/transcriptomic/day2_RNA_raw.csv"

# Loading data
day0_RNA_log2_FPKM <- as.data.frame(read_csv(day0_RNA_path))

day2_RNA_log2_FPKM <- as.data.frame(read_csv(day2_RNA_path))

# Change first column to rownames
day0_RNA_log2_FPKM <- tibble::column_to_rownames(day0_RNA_log2_FPKM, var = "...1")
day2_RNA_log2_FPKM <- tibble::column_to_rownames(day2_RNA_log2_FPKM, var = "...1")

# Remove redundant variables
rm(day0_RNA_path)
rm(day2_RNA_path)

```


### Plotting data

Day 0
```{r}
hist(as.numeric(unlist(day0_RNA_log2_FPKM)), breaks = 1000)
hist(as.numeric(unlist(day2_RNA_log2_FPKM)), breaks = 1000)

```
Many <1, so let's remove those and plot again:
```{r}
day0_RNA_log2_FPKM[day0_RNA_log2_FPKM < 1] <- NA
hist(as.numeric(unlist(day0_RNA_log2_FPKM)), breaks = 1000) 

day2_RNA_log2_FPKM[day2_RNA_log2_FPKM < 1] <- NA
hist(as.numeric(unlist(day2_RNA_log2_FPKM)), breaks = 1000) 
```

It does not work. I cry.

Let's remove all rows (genes) where >20% is NA (8 out of 10)

```{r}
day0_RNA_log2_FPKM <- day0_RNA_log2_FPKM[rowSums(is.na(day0_RNA_log2_FPKM)) < 8,]

hist(as.numeric(unlist(day0_RNA_log2_FPKM)), breaks = 1000) 


day2_RNA_log2_FPKM <- day2_RNA_log2_FPKM[rowSums(is.na(day2_RNA_log2_FPKM)) < 6,]

hist(as.numeric(unlist(day2_RNA_log2_FPKM)), breaks = 1000) 

```

