---
title: "3 - Preprocessing Metabolomics"
author: "Cathy Magnée, Saezlab"
date: "2024-06-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preprocessing metabolomics data 

This document describes the preprocessing of metabolomic T cell data. It is based off the preprocessing documents in the FACTOR_Cosmos No_MOFA branch by Saezlab on github. This document is preceded by the 2 - Preprocessing transcriptomic data R markdown document and precedes the 4 - Preparing COSMOS inputs R markdown document. 

This document is annotated to explain the steps and the choices that were made durnig this project. Other options will be highlighted. 

Document summary:
The main goal of this document is to transform the raw metabolite data into data that can be compared across sample. This will be done in the following steps and using the following tools:
1. Two steps of filtering (outliers)
2. Normalization using vsn normalization
3. Scaling using Z-score normalization.

#### Loading

Libraries
```{r, echo=FALSE, warning=FALSE}
library(readr)
library(reshape2)
library(pheatmap)
library(vsn)
library(tibble)
```

Data paths:
Splitting these in AMC and Laptop, so it is easier to switch. 

AMC:
```{r}
# T0_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T0_metabs_raw.csv"
# T48_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T48_metabs_raw.csv"

# or hmdb?
T0_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T0_metabs_hmdb.csv"
T48_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T48_metabs_hmdb.csv"

```
Laptop:
```{r}
# T0_path <- "../ourdata/metabolomic/T0_metabs_raw.csv"
# T48_path <- "../ourdata/metabolomic/T48_metabs_raw.csv"

# or hmdb?
# T0_path <- "../ourdata/metabolomic/T0_metabs_hmdb.csv"
# T48_path <- "../ourdata/metabolomic/T48_metabs_hmdb.csv"
```

Loading data and removing paths:
```{r}
# T0_metabs_raw <- as.data.frame(read_csv(T0_path))
# T48_metabs_raw <- as.data.frame(read_csv(T48_path))

# or hmdb?
T0_metabs_hmdb <- as.data.frame(read_csv(T0_path))
T48_metabs_hmdb <- as.data.frame(read_csv(T48_path))

# Adjusting first column to rownames
T0_metabs_hmdb <- tibble::column_to_rownames(T0_metabs_hmdb, var = "...1")
T48_metabs_hmdb <- tibble::column_to_rownames(T48_metabs_hmdb, var = "...1")

# Removing redundant paths
rm(T0_path)
rm(T48_path)

```

### Plotting and filtering

Plotting the data to see the distribution is the first step to understanding our data

```{r}
hist(as.numeric(unlist(log2(T0_metabs_hmdb))), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Log2 value")
mtext("Unstimulated CD3 T cells of HD and CLL patients", side = 3, line = 0.5)


hist(as.numeric(unlist(log2(T48_metabs_hmdb))), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Log2 value")
mtext("CD3 T cells of HD and CLL patients 48h after CD3 CD28 stimulation", side = 3, line = 0.5)


```
Now, let's also explore the data without log2 transformation:
```{r}
hist(as.numeric(unlist(T0_metabs_hmdb)), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Untransformed value")
mtext("Unstimulated CD3 T cells of HD and CLL patients", side = 3, line = 0.5)


hist(as.numeric(unlist(T48_metabs_hmdb)), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Untransformed value")
mtext("CD3 T cells of HD and CLL patients 48h after CD3 CD28 stimulation", side = 3, line = 0.5)
```
okay. Not great.Let's try filtering this data on very low values
```{r}
T0_metabs_hmdb[T0_metabs_hmdb < 0.00001] <- NA 
T48_metabs_hmdb[T48_metabs_hmdb < 0.00001] <- NA 

hist(as.numeric(unlist(T0_metabs_hmdb)), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Untransformed value")
mtext("Unstimulated CD3 T cells of HD and CLL patients", side = 3, line = 0.5)


hist(as.numeric(unlist(T48_metabs_hmdb)), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Untransformed value")
mtext("CD3 T cells of HD and CLL patients 48h after CD3 CD28 stimulation", side = 3, line = 0.5)
```

Now, transformed but per sample
T0:
```{r}
# Loop through each column to create histograms
par(mfrow = c(2, 5))
for (col_name in colnames(T0_metabs_hmdb)) {
  # Extract the current column and convert to log2 values
  current_col <- as.numeric(unlist(T0_metabs_hmdb[[col_name]]))
  
  # Create a histogram for the current column
  hist(current_col, breaks = 100,
       main = paste("Hist of metab val", col_name),
       xlab = "Untransformed value")
  
  # Add the subtitle below the main title
  # mtext("Unstimulated CD3 T cells of both HD and CLL patients", side = 3, line = 0.5)
  mtext("Unstim", side = 3, line = 0.5)

}
```
T48:
```{r}
par(mfrow = c(2, 4))
for (col_name in colnames(T48_metabs_hmdb)) {
  # Extract the current column and convert to log2 values
  current_col <- as.numeric(unlist(T48_metabs_hmdb[[col_name]]))
  
  # Create a histogram for the current column
  hist(current_col, breaks = 100,
       main = paste("Hist. metab values", col_name),
       xlab = "Untransformed value")
  
  # Add the subtitle below the main title
  # mtext("CD3 T cells of both HD and CLL patients 48h after CD3/CD28 stimulation", side = 3, line = 0.5)
  mtext("T48", side = 3, line = 0.6)
}
```
We want to know which metabolites are present in the lowest bins. Especially for CLL3045_T0 this can be interesting, as they have a lot of very low value outliers. 

```{r}
metabs_df  <- T0_metabs_hmdb
# metabs_df <- T48_metabs_hmdb

col_names <- colnames(metabs_df)

# Initialize a list to store metabolite names in the lowest value bins for each column
lowest_bins_metabolites <- list()

# Loop through each column name to analyze the histograms
for (col_name in col_names) {
  # Extract the current column by name and convert to log2 values
  current_col <- as.numeric(unlist(log2(metabs_df[[col_name]])))
  
  # Create a histogram for the current column without plotting it
  hist_data <- hist(current_col, breaks = 50, plot = FALSE)
  
  # Find the lowest value bin with non-zero counts
  lowest_bin <- which.min(hist_data$counts[hist_data$counts > 0])
  
  # Find the range of values in the lowest bin
  bin_range <- hist_data$breaks[c(lowest_bin, lowest_bin + 1)]
  
  # Identify metabolites in the lowest value bin
  metabolites_in_bin <- rownames(metabs_df)[log2(metabs_df[[col_name]]) >= bin_range[1]-100 & log2(metabs_df[[col_name]]) < bin_range[2]]
  
  # Store the results in the list
  lowest_bins_metabolites[[col_name]] <- metabolites_in_bin
}

# Print the list of metabolites in the lowest value bins
print(lowest_bins_metabolites)


```

```{r}
# Assuming T0_metabs_hmdb is a data frame with metabolite names as rownames and samples as columns
col_names <- colnames(T0_metabs_hmdb)

# Initialize a list to store metabolite names in the lowest value bins for each column
lowest_bins_metabolites <- list()

# Loop through each column name to analyze the histograms
for (col_name in col_names) {
  # Extract the current column by name and convert to log2 values, replacing -Inf with the minimum non -Inf value minus some epsilon
  current_col <- log2(T0_metabs_hmdb[[col_name]])
  current_col[is.infinite(current_col)] <- min(current_col[!is.infinite(current_col)]) - 1

  # Create a histogram for the current column without plotting it
  hist_data <- hist(current_col, breaks = 250, plot = FALSE)
  
  # Find the lowest value bin with non-zero counts
  lowest_bin <- which.min(hist_data$counts[hist_data$counts > 0])
  
  # Find the range of values in the lowest bin
  bin_range <- hist_data$breaks[c(lowest_bin, lowest_bin + 1)]
  
  # Identify metabolites in the lowest value bin or where the value is -Inf
  metabolites_in_bin <- rownames(T0_metabs_hmdb)[(current_col >= bin_range[1]-13 & current_col < bin_range[2]) | is.infinite(log2(T0_metabs_hmdb[[col_name]]))]
  
  # Store the results in the list
  lowest_bins_metabolites[[col_name]] <- metabolites_in_bin
}

# Print the list of metabolites in the lowest value bins
print(lowest_bins_metabolites)

```




Heatmaps:
```{r}
pheatmap(log2(T0_metabs_hmdb), show_colnames = F, show_rownames = F) 

# pheatmap(abs(log2(T48_metabs_hmdb)), show_colnames = F, show_rownames = F)

```


In the T48 dataset the histogram of the log2 values are all negative. This means all values are between 0 and 1. This should be adjusted for later wehen we use VSN normalization

The T0 heatmap also does not show much interesting as there are low values skewing the heatmap.

For the log2(T0) dataset, there are a some outliers <-30
For log2(T48), The outliers are -Inf (these are 0's)

Removing outliers:
```{r}
# T0_metabs_hmdb[log2(T0_metabs_hmdb) < -30] <- NA 
# T48_metabs_hmdb[log2(T48_metabs_hmdb) == -Inf] <- NA 

```

```{r}
hist(as.numeric(unlist(log2(T0_metabs_hmdb))), breaks = 1000)
hist(as.numeric(unlist(log2(T48_metabs_hmdb))), breaks = 1000)

```
Heatmaps after outlier removal:
```{r}
pheatmap(log2(T0_metabs_hmdb), show_colnames = F, show_rownames = F) 

pheatmap(abs(log2(T48_metabs_hmdb)), show_colnames = F, show_rownames = F)
```

Removing new outliers
- For log2 T0: <-1 and> 11
- For log2 T48: <-16 and >-6

```{r}
# T0_metabs_hmdb[log2(T0_metabs_hmdb) < -1] <- NA 
# T0_metabs_hmdb[log2(T0_metabs_hmdb) > 11 ] <- NA 
# 
# T48_metabs_hmdb[log2(T48_metabs_hmdb) < -16] <- NA 
# T48_metabs_hmdb[log2(T48_metabs_hmdb) > 6] <- NA 
```


The histogram values of (log) T48 are only negative. This means the m/z ratios are all between 0-1. 

VSN normalization is used to adjust values to be able to compare across samples


```{r}
hist(as.numeric(unlist(log2(T0_metabs_hmdb))), breaks = 1000)
hist(as.numeric(unlist(log2(T48_metabs_hmdb))), breaks = 1000)

```

### VSN normalization
First for T0
```{r}
# T0_metabs_hmdb <- T0_metabs_hmdb[,-4]
fit <- vsnMatrix(as.matrix(T0_metabs_hmdb))
meanSdPlot(fit)
T0_metabs_hmdb <- as.data.frame(vsn::predict(fit,as.matrix(T0_metabs_hmdb)))

```
Note that in the above chunk, CLL2474 is removed from the dataset as it proved to be an outlier in the downstream analysis. I do not want this to influence the vsnMatrix normalization. 

Then for T48:
```{r}
fit <- vsnMatrix(as.matrix(T48_metabs_hmdb))
meanSdPlot(fit)
T48_metabs_hmdb <- as.data.frame(vsn::predict(fit,as.matrix(T48_metabs_hmdb)))
```

```{r}
pheatmap(T0_metabs_hmdb[complete.cases(T0_metabs_hmdb),], show_colnames = T, show_rownames = F, cluster_rows = F) # , color = colorRampPalette(c("lightgrey","cornflowerblue", "blue"))(100)

pheatmap(T48_metabs_hmdb[complete.cases(T48_metabs_hmdb),], show_colnames = T, show_rownames = F, cluster_rows = F) # , color = colorRampPalette(c("forestgreen","lightgrey", "blue"))(100))  
```

### Scaling using Z-normalization
T0:
```{r}

SDs <- apply(T0_metabs_hmdb,1,function(x){sd(x,na.rm = T)})
means <- rowMeans(T0_metabs_hmdb, na.rm = T)
T0_metabs_scaled <- (T0_metabs_hmdb - means) / SDs
pheatmap(T0_metabs_scaled[complete.cases(T0_metabs_scaled),], show_colnames = T, show_rownames = F, cluster_rows = F) 

hist(as.numeric(unlist(T0_metabs_scaled)), breaks = 1000) 
```

CLL2474 is an outlier. For now, this one is taken out (this has been done previously)

T48:
```{r}
SDs <- apply(T48_metabs_hmdb,1,function(x){sd(x,na.rm = T)})
means <- rowMeans(T48_metabs_hmdb, na.rm = T)
T48_metabs_scaled <- (T48_metabs_hmdb - means) / SDs
pheatmap(T48_metabs_scaled[complete.cases(T48_metabs_scaled),], show_colnames = T, show_rownames = F, cluster_rows = F) 

hist(as.numeric(unlist(T48_metabs_scaled)), breaks = 1000) 
```


### Save data
AMC
```{r}
############ Paths to write data to AMC PC ############ 
write.csv(T0_metabs_scaled,file = "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T0_metabs_hmdb_scaled.csv")
write.csv(T48_metabs_scaled,file = "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T48_metabs_hmdb_scaled.csv")
```

Laptop
```{r}
############ Paths to write data to laptop ############ 
# write.csv(T0_metabs_scaled,file = "../ourdata/metabolomic/T0_metabs_hmdb_scaled.csv")
# write.csv(T48_metabs_scaled,file = "../ourdata/metabolomic/T48_metabs_hmdb_scaled.csv")
```


Explore again per sample:
```{r}
par(mfrow = c(2, 5))
for (col_name in colnames(T0_metabs_scaled)) {
  # Extract the current column and convert to log2 values
  current_col <- as.numeric(unlist(T0_metabs_scaled[[col_name]]))
  
  # Create a histogram for the current column
  hist(current_col, breaks = 100,
       main = paste("Hist of metab val", col_name),
       xlab = "Transformed values")
  
  # Add the subtitle below the main title
  # mtext("Unstimulated CD3 T cells of both HD and CLL patients", side = 3, line = 0.5)
  mtext("Unstim", side = 3, line = 0.5)

}
```
```{r}
par(mfrow = c(2, 4))
for (col_name in colnames(T48_metabs_scaled)) {
  # Extract the current column and convert to log2 values
  current_col <- as.numeric(unlist(T48_metabs_scaled[[col_name]]))
  
  # Create a histogram for the current column
  hist(current_col, breaks = 100,
       main = paste("Hist of metab val", col_name),
       xlab = "Transformed values")
  
  # Add the subtitle below the main title
  # mtext("Unstimulated CD3 T cells of both HD and CLL patients", side = 3, line = 0.5)
  mtext("T48", side = 3, line = 0.5)

}

```

