---
title: "3 - Preprocessing Metabolomics"
author: "Cathy Magnée, Saezlab"
date: "2024-06-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preprocessing metabolomics data 

This document describes the preprocessing of metabolomic CD3 T cell data. It is based off the preprocessing documents in the FACTOR_Cosmos No_MOFA branch by Saezlab on github. This document is preceded by the 2 - Preprocessing transcriptomic data R markdown document and precedes the 4 - Preparing COSMOS inputs R markdown document. 

This document is annotated to explain the steps and the choices that were made during this project. Other options will be highlighted. 

Document summary:
The main goal of this document is to transform the raw metabolite data into data that can be compared across samples and the sample groups. This will be done in the following steps and using the following tools:
1. Filtering outliers
2. Normalization using VSN normalization
3. Scaling using Z-score normalization.

#### Loading

Libraries
```{r, echo=FALSE, warning=FALSE}
library(readr)
library(reshape2)
library(pheatmap)
library(vsn)
library(tibble)
library(RColorBrewer)
```

Data paths:
Splitting these in AMC and Laptop, so it is easier to switch. 

AMC paths:
```{r}
# Metabolite names as rownames
# T0_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T0_metabs_raw.csv"
# T48_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T48_metabs_raw.csv"

# HMDB IDs as rownames
T0_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T0_metabs_hmdb.csv"
T48_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T48_metabs_hmdb.csv"

```

Laptop paths:
```{r}
# Metabolite names as rownames
# T0_path <- "../ourdata/metabolomic/T0_metabs_raw.csv"
# T48_path <- "../ourdata/metabolomic/T48_metabs_raw.csv"

# HMDB IDs as rownames
# T0_path <- "../ourdata/metabolomic/T0_metabs_hmdb.csv"
# T48_path <- "../ourdata/metabolomic/T48_metabs_hmdb.csv"
```

Loading data and removing path variables:
```{r, message=FALSE}
# Metabolite names as rownames
# T0_metabs_raw <- as.data.frame(read_csv(T0_path))
# T48_metabs_raw <- as.data.frame(read_csv(T48_path))
# T0_metabs_hmdb <- T0_metabs_raw
# T48_metabs_hmdb <- T48_metabs_raw

# HMDB IDs as rownames
T0_metabs_hmdb <- as.data.frame(read_csv(T0_path, show_col_types = FALSE))
T48_metabs_hmdb <- as.data.frame(read_csv(T48_path, show_col_types = FALSE))

# First column -> Rownames, then remove first column
T0_metabs_hmdb <- tibble::column_to_rownames(T0_metabs_hmdb, var = "...1")
T48_metabs_hmdb <- tibble::column_to_rownames(T48_metabs_hmdb, var = "...1")

# Removing now-redundant path variables
rm(T0_path)
rm(T48_path)

```

### Plotting and filtering

Plotting the data to see the distribution is the first step to understanding our data. Let's first see the histograms. 

```{r}
hist(as.numeric(unlist(log2(T0_metabs_hmdb))), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Log2 value")
mtext("Unstimulated CD3 T cells of HD and CLL patients", side = 3, line = 0.5)


hist(as.numeric(unlist(log2(T48_metabs_hmdb))), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Log2 value")
mtext("CD3 T cells of HD and CLL patients 48h after CD3 CD28 stimulation", side = 3, line = 0.5)


```
There are some outliers in the left-most bins. This is especially notable in the histogram representing the log-transformed values in the unstimulated dataset. What metabolites are causing these left-most bins and why are the values so low?
Let's investigate this in the following chunk, in which we check the lowest-most bin:

```{r}
# Choose here which timepoint to analyze
metabs_df <- T48_metabs_hmdb

# Apply log2 transformation safely
log_metabs_df <- log2(metabs_df)

# Create a histogram for the entire dataset (flattened into a single vector)
hist_data <- hist(as.numeric(unlist(log_metabs_df)), breaks = 250, plot = FALSE)

# Find the lowest value bin with non-zero counts
lowest_bin <- min(which(hist_data$counts > 0))

# Find the range of values in the lowest bin
bin_range <- hist_data$breaks[c(lowest_bin, lowest_bin + 1)]

# Debugging: Print the range of the lowest bin
print(paste("Lowest bin range:", bin_range[1], "to", bin_range[2]))

# Identify metabolites in the lowest value bin
metabolites_in_bin <- rownames(metabs_df)[apply(log_metabs_df, 1, function(row) any(row >= bin_range[1] & row < bin_range[2]))]

# Debugging: Check if NA values exist in the resulting list
if (any(is.na(metabolites_in_bin))) {
  warning("NA values found in the metabolites_in_bin list.")
}

# Print the list of metabolites in the lowest value bins
print(metabolites_in_bin)

```
Let's remove the (very low value) outliers and explore the dataset again:
```{r}
T0_metabs_hmdb[T0_metabs_hmdb < 0.00001] <- NA 
T48_metabs_hmdb[T48_metabs_hmdb < 0.00001] <- NA 

hist(as.numeric(unlist(log2(T0_metabs_hmdb))), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Log2 value")
mtext("Unstimulated CD3 T cells of HD and CLL patients", side = 3, line = 0.5)
mtext("After outlier removal", side = 3, line = -0.5)



hist(as.numeric(unlist(log2(T48_metabs_hmdb))), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Log2 value")
mtext("CD3 T cells of HD and CLL patients 48h after CD3 CD28 stimulation", side = 3, line = 0.5)
mtext("After outlier removal", side = 3, line = -0.5)


```
This looks better! 
Now, let's also explore the data without log2 transformation:
```{r}
hist(as.numeric(unlist(T0_metabs_hmdb)), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Untransformed value")
mtext("Unstimulated CD3 T cells of HD and CLL patients", side = 3, line = 0.5)


hist(as.numeric(unlist(T48_metabs_hmdb)), breaks = 250,
     main = "Histogram of metabolite values",
     xlab = "Untransformed value")
mtext("CD3 T cells of HD and CLL patients 48h after CD3 CD28 stimulation", side = 3, line = 0.5)
```
These two figures are hard to interpret. Let's explore this data per sample
Okay. Not great. Let's try gain after filtering this data on the very low values (outliers).


```{r}
metabs_df <- T0_metabs_hmdb
par(mfrow = c(2, 5))

# metabs_df <- T48_metabs_hmdb
# par(mfrow = c(2, 4))

# Loop through each column to create histograms

for (col_name in colnames(metabs_df)) {
  # Extract the current column and convert to log2 values
  current_col <- as.numeric(unlist(metabs_df[[col_name]]))
  
  # Create a histogram for the current column
  hist(current_col, breaks = 100,
       main = paste("Hist of metab val", col_name),
       xlab = "Untransformed value")
  
  # Add the subtitle below the main title
  # mtext("Unstimulated CD3 T cells of both HD and CLL patients", side = 3, line = 0.5)
  # mtext("T0", side = 3, line = 0.5)
  # mtext("T48", side = 3, line = 0.5)

}
par(mfrow = c(1, 1))
```

Let's analyze for each sample what metabolites are present in the lowest (transformed) bins:

```{r}
# Assuming T0_metabs_hmdb is a data frame with metabolite names as rownames and samples as columns
col_names <- colnames(T0_metabs_hmdb)

# Initialize a list to store metabolite names in the lowest value bins for each column
lowest_bins_metabolites <- list()

# Loop through each column name to analyze the histograms
for (col_name in col_names) {
  # Extract the current column by name and convert to log2 values, replacing -Inf with the minimum non -Inf value minus some epsilon
  current_col <- log2(T0_metabs_hmdb[[col_name]])
  current_col[is.infinite(current_col)] <- min(current_col[!is.infinite(current_col)]) - 1

  # Create a histogram for the current column without plotting it
  hist_data <- hist(current_col, breaks = 250, plot = FALSE)
  
  # Find the lowest value bin with non-zero counts
  lowest_bin <- which.min(hist_data$counts[hist_data$counts > 0])
  
  # Find the range of values in the lowest bin
  bin_range <- hist_data$breaks[c(lowest_bin, lowest_bin + 1)]
  
  # Identify metabolites in the lowest value bin or where the value is -Inf
  metabolites_in_bin <- rownames(T0_metabs_hmdb)[(current_col >= bin_range[1]-13 & current_col < bin_range[2]) | is.infinite(log2(T0_metabs_hmdb[[col_name]]))]
  
  # Store the results in the list
  lowest_bins_metabolites[[col_name]] <- metabolites_in_bin
}

# Print the list of metabolites in the lowest value bins
# View(lowest_bins_metabolites)

```

These are often the same metabolite. 
As the individual plots show, there are often 'outlier' values in the right-most part of the plot. Let's find out which metabolites cause this right-most skewing. 
In the chunk below, we try to find the metabolites in the five right-most histogram bins of untransformed data per sample.
```{r}
DF <- T0_metabs_hmdb
# DF <- T48_metabs_hmdb
col_names <- colnames(DF)
# Initialize a list to store metabolite names in the 5 right-most bins for each column
rightmost_bins_metabolites <- list()

# Loop through each column name to analyze the histograms
for (col_name in col_names) {
  # Extract the current column by name (original values, no log2)
  current_col <- as.numeric(DF[[col_name]])

  # Create a histogram for the current column without plotting it
  hist_data <- hist(current_col, breaks = 50, plot = FALSE)
  
  # Find the indices of the non-empty bins
  non_empty_bins <- which(hist_data$counts > 0)
  
  # Determine the range of bins to consider (the 5 right-most non-empty bins)
  bins_to_consider <- tail(non_empty_bins, 5)
  
  # Initialize a list to store metabolites for the current column
  metabolites_list <- list()

  # Loop through the 5 right-most bins
  for (bin_index in bins_to_consider) {
    # Find the range of values in the current bin
    bin_range <- hist_data$breaks[c(bin_index, bin_index + 1)]
    
    # Identify metabolites in the current bin
    metabolites_in_bin <- rownames(DF)[(current_col >= bin_range[1] & current_col < bin_range[2])]
    
    # Store the metabolites in the list for the current bin
    metabolites_list[[paste0("Bin_", bin_index)]] <- metabolites_in_bin
  }
  
  # Store the results for the current column in the overall list
  rightmost_bins_metabolites[[col_name]] <- metabolites_list
}

# Print the list of metabolites in the 5 right-most non-empty bins
# View(rightmost_bins_metabolites)

```

Let's now explore our data further.

Heatmaps:
```{r}
pheatmap(log2(T0_metabs_hmdb), show_colnames = T, show_rownames = F, color = hcl.colors(50, "Oslo")) 

pheatmap(abs(log2(T48_metabs_hmdb)), show_colnames = T, show_rownames = F, color = hcl.colors(50, "Oslo")) 

```

Re-explore histogram (after outlier removal):
```{r}
hist(as.numeric(unlist(log2(T0_metabs_hmdb))), breaks = 200)
hist(as.numeric(unlist(log2(T48_metabs_hmdb))), breaks = 200)
```


The histogram values of the log2-tranformed T48 dataset are all negative. This means the m/z ratio values are all between 0-1. 
VSN normalization is used to adjust values to be able to compare across samples.

### VSN normalization
First for T0
```{r}
T0_metabs_hmdb <- T0_metabs_hmdb[, !(colnames(T0_metabs_hmdb) %in% "CLL2474_T0")]
fit <- vsnMatrix(as.matrix(T0_metabs_hmdb))
meanSdPlot(fit)
T0_metabs_hmdb <- as.data.frame(vsn::predict(fit,as.matrix(T0_metabs_hmdb)))

```

Note that in the above chunk, CLL2474 is removed from the dataset as it proved to be an outlier in the downstream analysis. I do not want this to influence the vsnMatrix normalization. 

Then for T48:
```{r}
fit <- vsnMatrix(as.matrix(T48_metabs_hmdb))
meanSdPlot(fit)
T48_metabs_hmdb <- as.data.frame(vsn::predict(fit,as.matrix(T48_metabs_hmdb)))
```

```{r}
pheatmap(T0_metabs_hmdb[complete.cases(T0_metabs_hmdb),], show_colnames = T, show_rownames = F, cluster_rows = F , color = hcl.colors(50, "Oslo")) 

pheatmap(T48_metabs_hmdb[complete.cases(T48_metabs_hmdb),], show_colnames = T, show_rownames = F, cluster_rows = F, color = hcl.colors(50, "Oslo"))  
```

### Scaling using Z-normalization
T0:
```{r}
# Scaling using Z-normalization:
SDs <- apply(T0_metabs_hmdb,1,function(x){sd(x,na.rm = T)})
means <- rowMeans(T0_metabs_hmdb, na.rm = T)
T0_metabs_scaled <- (T0_metabs_hmdb - means) / SDs

# Data exploration
pheatmap(T0_metabs_scaled[complete.cases(T0_metabs_scaled),], show_colnames = T, show_rownames = F, cluster_rows = F, color = hcl.colors(50, "Oslo"))

hist(as.numeric(unlist(T0_metabs_scaled)), breaks = 250, 
     main = "Histogram of unstim CD3 T cell metabolite values",
     xlab = "Transformed m/z ratio values")
mtext("Unstimulated CD3 T cells of HD and CLL patients", side = 3, line = 0.5)
mtext("After outlier removal, VSNmatrix normalization and Z-score normalization", side = 3, line = -0.5) 
```

CLL2474 is an outlier. For now, this one is taken out (this has been done previously)

T48:
```{r}
# Scaling using Z-normalization:
SDs <- apply(T48_metabs_hmdb,1,function(x){sd(x,na.rm = T)})
means <- rowMeans(T48_metabs_hmdb, na.rm = T)
T48_metabs_scaled <- (T48_metabs_hmdb - means) / SDs

# Data exploration
pheatmap(T48_metabs_scaled[complete.cases(T48_metabs_scaled),], show_colnames = T, show_rownames = F, cluster_rows = F, color = hcl.colors(50, "Oslo"))

hist(as.numeric(unlist(T48_metabs_scaled)), breaks = 250, 
     main = "Histogram of stimulated CD3 T cell metabolite values",
     xlab = "Transformed m/z ratio values")
mtext("CD3 T cells of HD and CLL patients, 48h after CD3 CD28 stimulation", side = 3, line = 0.5)
mtext("After outlier removal, VSNmatrix normalization and Z-score normalization", side = 3, line = -0.5) 
```


### Save data
AMC
```{r}
############ Paths to write data to AMC PC ############ 
write.csv(T0_metabs_scaled,file = "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T0_metabs_hmdb_scaled.csv")
write.csv(T48_metabs_scaled,file = "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T48_metabs_hmdb_scaled.csv")
```

Laptop
```{r}
############ Paths to write data to laptop ############ 
# write.csv(T0_metabs_scaled,file = "../ourdata/metabolomic/T0_metabs_hmdb_scaled.csv")
# write.csv(T48_metabs_scaled,file = "../ourdata/metabolomic/T48_metabs_hmdb_scaled.csv")
```


In the chunks below, the data distribution is visualized again, per sample, via histograms: :
```{r}
metab_df <- T0_metabs_scaled
par(mfrow = c(2, 5))

# metab_df <- T48_metabs_scaled
# par(mfrow = c(2, 4))

for (col_name in colnames(metab_df)) {
  # Extract the current column and convert to log2 values
  current_col <- as.numeric(unlist(metab_df[[col_name]]))
  
  # Create a histogram for the current column
  hist(current_col, breaks = 100,
       main = paste("Hist of metab val", col_name),
       xlab = "Transformed values")
  
  # Add the subtitle below the main title
  # mtext("Unstimulated CD3 T cells of both HD and CLL patients", side = 3, line = 0.5)
  mtext("Unstim", side = 3, line = 0.5)

}
```


