---
title: "4 - Preparing COSMOS inputs"
author: "Cathy Magnée, Saezlab"
date: "2024-06-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preparing COSMOS inputs

This document describes the process of preparing metabolite and transcriptomic T cell data to form the input of COSMOS. This requires two main steps:
1. Estimating TF activities from T cell transcriptomic data using DecoupleR and CollecTRI
2. Combining the three datasets into one dataframe

This script is based off the preprocessing documents in the FACTOR_Cosmos No_MOFA branch by Saezlab on github. This document is preceded by the 3 - Preprocessing gene expression data R markdown document and precedes the 5 - Run COSMOS R markdown document. 

This document is annotated to explain the steps and the choices that were made during this project. Other options will be highlighted.

Document summary:
The aim of this document is both finding a suitable method to apply to the gene expression data, using this to estimate TF activities and combining the three datasets (RNA, signalling and metabolite data) into cosmos_inputs dataset. This will be done side-by-side for two datasets, unstimulated (T0) and 48 hours after stimulation (T48), each consisting of healthy donors (HD) and CLL patients data. 

All this will happen in X steps:
1. Make data uniform (Exclude non-overlapping columns)
2. Test methods for TF activity estimation
3. Perform TF activity estimation
4. Reformatting data into cosmos_inputs data file. 

#### Loading
In case you need it: Install decoupleR here:
```{r, echo = FALSE, warning=FALSE}
# install.packages('remotes')
# remotes::install_github('saezlab/decoupleR')
# remotes::install_github('saezlab/OmnipathR')

## To get Omnipath to work, I had to remove package xfun and install it via the cran webpage. I did this via:
# packageurl <- "https://cran.r-project.org/bin/windows/contrib/4.5/xfun_0.45.zip"
# > install.packages(packageurl, repos=NULL, type="source")

```


DecoupleR source:
[1] Badia-i-Mompel P., Vélez Santiago J., Braunger J., Geiss C., Dimitrov D., Müller-Dott S., Taus P., Dugourd A., Holland C.H., Ramirez Flores R.O. and Saez-Rodriguez J. 2022. decoupleR: ensemble of computational methods to infer biological activities from omics data. Bioinformatics Advances. <https://doi.org/10.1093/bioadv/vbac016>

When loading we will also load CollecTRI. 
CollecTRI research source:
[2]Müller-Dott S., Tsirvouli E., Vázquez M., Ramirez Flores R.O., Badia-i-Mompel P., Fallegger R., Lægreid A. and Saez-Rodriguez J. Expanding the coverage of regulons from high-confidence prior knowledge for accurate estimation of transcription factor activities. bioRxiv. 2023. DOI: <10.1101/2023.03.30.534849> 

CollecTRI data source:

Loading Libraries
```{r}
library(readr)
library(decoupleR)
library(OmnipathR)
library(dplyr)
```

Writing down data paths. This goes in two separate chunks: One for my AMC pc and one for my laptop. This way, it is easier to switch between the two. 
Note that at the bottom of this document, there will also be two chunks to switch between based on workstation for saving the end result of the script. 

AMC:
```{r}
############ Paths to load data from AMC PC ############ 
# T0
path_T0_metabolomic_data <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T0_metabs_hmdb_scaled.csv"
path_T0_transcriptomic_data <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Transcriptomic\\T0_RNA_log2_FPKM_cleaned.csv"

#T48
path_T48_metabolomic_data <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Metabolomic\\T48_metabs_hmdb_scaled.csv"
path_T48_transcriptomic_data <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\Transcriptomic\\T48_RNA_log2_FPKM_cleaned.csv"


```

Laptop:
```{r}
############ Paths to load data from Laptop ############ 
# T0
# path_T0_metabolomic_data <- "../ourdata/metabolomic/T0_metabs_hmdb_scaled.csv"
# path_T0_transcriptomic_data <- "../ourdata/transcriptomic/T0_RNA_log2_FPKM_cleaned.csv"
# 
# # T48
# path_T48_metabolomic_data <-  "../ourdata/metabolomic/T48_metabs_hmdb_scaled.csv"
# path_T48_transcriptomic_data <- "../ourdata/transcriptomic/T48_RNA_log2_FPKM_cleaned.csv"

```

Loading data and removing paths:
```{r, message=FALSE, warning=FALSE}
T0_metabolomic <- as.data.frame(read_csv(path_T0_metabolomic_data, show_col_types = FALSE))
names(T0_metabolomic)[names(T0_metabolomic) == "...1"] <- "metabolite"

T0_transcriptomic <- as.data.frame(read_csv(path_T0_transcriptomic_data, show_col_types = FALSE))
rownames(T0_transcriptomic) <- T0_transcriptomic$...1

T48_metabolomic <- as.data.frame(read_csv(path_T48_metabolomic_data, show_col_types = FALSE))
names(T48_metabolomic)[names(T48_metabolomic) == "...1"] <- "metabolite"

T48_transcriptomic <-  as.data.frame(read_csv(path_T48_transcriptomic_data, show_col_types = FALSE))
rownames(T48_transcriptomic) <- T48_transcriptomic$...1

# CollecTRI <- decoupleR::get_collectri(organism='human', split_complexes=FALSE)

# Remove redundant variables (paths)
rm(path_T0_metabolomic_data) 
rm(path_T0_transcriptomic_data)
rm(path_T48_metabolomic_data) 
rm(path_T48_transcriptomic_data)
```

### Intersect dataframes

First, we want to make sure we have a complete gene expression and metabolomic dataset for each patient / donor. 
We will test this by intersecting the column names of both dataframes and using only these column names for both.
T0:
```{r}
# intersect so only HD and CLL patients available in both datasets are used
samples <- intersect(names(T0_metabolomic),names(T0_transcriptomic)) 

T0_metabolomic <- T0_metabolomic[,c("metabolite",samples)]
T0_transcriptomic <- T0_transcriptomic[,samples]
```

T48:
```{r}
# intersect so only HD and CLL patients available in both datasets are used
samples <- intersect(names(T48_metabolomic),names(T48_transcriptomic))

T48_metabolomic <- T48_metabolomic[,c("metabolite",samples)]
T48_transcriptomic <- T48_transcriptomic[,samples]
```

## TF Activity Estimation
To estimate TF activity, regulons are needed. Regulons are the TF and their gene expression targets (so what targets are targeted by the TF). For this, the CollecTRI database is the most recent and recommended by Saezlab. 

### Regulon loading and cleaning
Loading CollecTRI processed regulons from Omnipath
```{r, message=FALSE, warning= FALSE}
# Omnipath settings
collectri_regulons <- get_collectri(organism = 'human', split_complexes = FALSE)
collectri_regulons <- decoupleR::get_collectri()
collectri_regulons <- collectri_regulons[!(is.na(collectri_regulons[, 2]) | collectri_regulons[, 2] == ""), ]
```
For some reason, using the above df does not work. The one below does, but it needs more preprocessing. Thus, the following two chunks are the importing and preprocessing of these regulons. In the end, we use both datasets.

Same CollecTRI regulons, different route:
```{r, message=FALSE, warning= FALSE}

# Omnipath settings to get collectri regulons

omnipath_set_cachedir('tmpcache')
omnipath_set_console_loglevel('trace')

# Getting all regulons
ci <- collectri() # overview

```

This db consists of much more info. Some sources and targets have an stimulating and inhibiting role. For these, the collectri_regulons from the first chunk will be used as a reference. If this dataset shows stimulating (1) this will be copied, and for inhibiting (-1) this will be translated to a 0 (non-activating)

The 0's are then later all changed to -1.

Cleaning up the CollecTRI regulons:
```{r, message=FALSE}

ci_regulons <- ci[,c(3,4,6)] # only need these columns from overview. These are source, target and stimulates (0 and 1)

colnames(ci_regulons) <- c("source", "target", "mor")
ci_regulons <- unique(ci_regulons) # remove pure dupes

# Some of the regulons are duplicated with different mor. 
# This means the same source/target combination has both 0 and 1 options
# Let's change this. 
temp <- ci[duplicated(ci[, 3:4]) | duplicated(ci[, 3:4], fromLast = TRUE), ] # all duplicates, also with the same mor values
duplicates <- anti_join(temp, ci[duplicated(ci[, c(3,4,6)]) | duplicated(ci[, c(3,4,6)], fromLast = TRUE), ]) # only different mor values
duplicates <- duplicates[,c(3,4,6)]
colnames(duplicates) <- c("source", "target", "mor")
duplicates <- unique(duplicates)

# For loop that checks the value of the duplicates in ci_regulons what their values are in collectri_regulons then changes accordingly 
for(i in 1:nrow(duplicates)){
  source <- duplicates$source[i]
  target <- duplicates$target[i]
  
  collectri_row <- collectri_regulons %>% 
    filter(source == !!source & target == !!target)
  
  if (nrow(collectri_row) > 0){ # if it is in collectri
    # check mor value
    if (collectri_row$mor == -1){
      ci_regulons <- ci_regulons |>
        mutate(mor = ifelse(source == !!source & target == !!target, 0 , mor))
    } else {
      ci_regulons <- ci_regulons |>
        mutate(mor = ifelse(source == !!source & target == !!target, 1, mor))
    }
  }
}

ci_regulons <- unique(ci_regulons)


# Change inhibition relation to -1 
ci_regulons$mor[ci_regulons$mor == 0] <- -1

```

Estimating TF activities.
There are multiple methods available to estimate TF activities. In the original pipeline (Factor cosmos no mofa), the run_wmean function is used. The official source of DecoupleR [1] describes the following three as most consistently better-performing: consensus, multivariate linear model and ULM. 

First, three methods are explored in the following three chunks. Then, the main method, which can be adjusted, is separately applied to the T0 and T48 gene expression datasets to estimate TF activity.

### Method exploration
For now, methods wmean, Ulm, mlm and consensus are explored.
Each is tested using the T0 gene expression dataset. This can be easily adjusted to be any other (preprocessed) dataset. 

Wmean
```{r}
# RNA_scaled <- T0_transcriptomic
# 
# T0_TF_activities <- apply(RNA_scaled,2,function(x){
#                         x <- as.data.frame(x[which(!is.na(x))])
#                         TFs <- run_wmean(as.matrix(x),
#                                          network = unique(ci_regulons),
#                                          times = 1000,
#                                          minsize = 20)
#                         TFs <- as.data.frame(TFs)
#                         TFs <- TFs[which(TFs$statistic == "norm_wmean"),c(2,4)]
#                         as_input <- TFs[,2]
#                         names(as_input) <- TFs[,1]
#                         return(as_input)
#                         })
```

ULM
```{r}
# RNA_scaled <- T0_transcriptomic
# 
# T0_TF_activities <- apply(RNA_scaled,2,function(x){
#                         x <- as.data.frame(x[which(!is.na(x))]) # no NAs allowed
#                         TFs <- run_ulm(as.matrix(x), # actual running of TF activity estimation function
#                                          network = unique(ci_regulons),
#                                          # times = 1000,
#                                          minsize = 20)
#                         TFs <- as.data.frame(TFs) # Reformatting into cosmos input format
#                         print(head(TFs))
#                         TFs <- TFs[which(TFs$statistic == "ulm"),c(2,4)]
#                         as_input <- TFs[,2]
#                         names(as_input) <- TFs[,1]
#                         return(as_input)
#                         })
```


Consensus
Consensus and MLM both have a two-step approach, as it does not allow for colinear regulons, these will be filtered out of the PKN before using it as input. 
1. Adjusting PKN
```{r}

# correlated_regulons <- decoupleR::check_corr(ci_regulons)
# 
# correlated_vars <- correlated_regulons |>
#   filter(correlation > 0.5 | correlation < -0.5  ) |>
#   dplyr::select(source, source.2) |>
#   unlist()
# 
# ci_regulons_no_cor <- ci_regulons |>
#   filter(!(source %in% correlated_vars | target %in% correlated_vars))
```

2. Combining new PKN and gene expression data to estimate TF activities
```{r}
# RNA_scaled <- T0_transcriptomic
# 
# T0_TF_activities <- apply(RNA_scaled,2,function(x){
#   x <- as.data.frame(x[which(!is.na(x))])
#   TFs <- decouple(as.matrix(x), # actual running of TF activity estimation function
#                    network = unique(ci_regulons_no_cor), 
#                   # statistics = c("ulm", "mlm", "wmean"),
#                   # args = list(
#                   #   ulm = list(.mor = "mor", .likelihood = "likelihood"),
#                   #   mlm = list(.mor = "mor", .likelihood = "likelihood"),
#                   #   wmean = list(.mor = "mor", .likelihood = "likelihood"))
#                   #   ),
#                    minsize = 20)
#   TFS <- run_consensus(TFs)
#   # print(head(TFs))
#   TFs <- as.data.frame(TFs) # Reformatting into cosmos input format
#   TFs <- TFs[which(TFs$statistic == "consensus"),c(3,5)]
#   # print(head(TFs))
#   as_input <- TFs[,2]
#   # print(head(as_input))
#   names(as_input) <- TFs[,1]
#   # print(head(as_input))
#   return(as_input)
#   })
```

Note that it is also possible to un-comment the statistics and args arguments in the decouple function to edit the statistical methods that are being weighted in the consensus function. By using the default values, we are using the methods mlm, ulm and wsum. 

Overview available methods:
<https://decoupler-py.readthedocs.io/en/latest/notebooks/usage.html>

MLM
```{r}
# RNA_scaled <- T0_transcriptomic
# 
# T0_TF_activities <- apply(RNA_scaled,2,function(x){
#                         x <- as.data.frame(x[which(!is.na(x))]) # no NAs allowed
#                         TFs <- run_mlm(as.matrix(x), # actual running of TF activity estimation function
#                                          network = unique(ci_regulons_no_cor),
#                                          # times = 1000,
#                                          minsize = 20)
#                         TFs <- as.data.frame(TFs) # Reformatting into cosmos input format
#                         TFs <- TFs[which(TFs$statistic == "mlm"),c(2,4)]
#                         as_input <- TFs[,2]
#                         names(as_input) <- TFs[,1]
#                         return(as_input)
#                         })
```


### Applying method
The consensus method will be applied as this one is described in literature to be the most consistent and best performing [1].

Preprocessing step:
```{r}
# correlated_regulons <- decoupleR::check_corr(ci_regulons)
# 
# correlated_vars <- correlated_regulons |>
#   filter(correlation > 0.5 | correlation < -0.5  ) |>
#   dplyr::select(source, source.2) |>
#   unlist()
# 
# ci_regulons_no_cor <- ci_regulons |>
#   filter(!(source %in% correlated_vars | target %in% correlated_vars))
```

T0:
```{r}
# RNA_scaled <- T0_transcriptomic
# 
# T0_TF_activities <- apply(RNA_scaled,2,function(x){
#   x <- as.data.frame(x[which(!is.na(x))])
#   TFs <- decouple(as.matrix(x), # actual running of TF activity estimation function
#                    network = unique(ci_regulons_no_cor),
#                    # times = 1000,
#                    minsize = 20)
#   TFS <- run_consensus(TFs)
#   # print(head(TFs))
#   TFs <- as.data.frame(TFs) # Reformatting into cosmos input format
#   TFs <- TFs[which(TFs$statistic == "consensus"),c(3,5)]
#   # print(head(TFs))
#   as_input <- TFs[,2]
#   # print(head(as_input))
#   names(as_input) <- TFs[,1]
#   # print(head(as_input))
#   return(as_input)
# })

```

T48:
```{r}
# RNA_scaled <- T48_transcriptomic
# 
# T48_TF_activities <- apply(RNA_scaled,2,function(x){
#   x <- as.data.frame(x[which(!is.na(x))])
#   TFs <- decouple(as.matrix(x), # actual running of TF activity estimation function
#                    network = unique(ci_regulons_no_cor),
#                    # times = 1000,
#                    minsize = 20)
#   TFS <- run_consensus(TFs)
#   # print(head(TFs))
#   TFs <- as.data.frame(TFs) # Reformatting into cosmos input format
#   TFs <- TFs[which(TFs$statistic == "consensus"),c(3,5)]
#   # print(head(TFs))
#   as_input <- TFs[,2]
#   # print(head(as_input))
#   names(as_input) <- TFs[,1]
#   # print(head(as_input))
#   return(as_input)
  # })
```

Wmean is used in the Factor_no COSMOS pipeline, so I want to try that too:
T0:
```{r}
# RNA_scaled <- T0_transcriptomic
# 
# T0_TF_activities <- apply(RNA_scaled,2,function(x){
#                         x <- as.data.frame(x[which(!is.na(x))])
#                         TFs <- run_wmean(as.matrix(x),
#                                          network = unique(ci_regulons),
#                                          times = 1000,
#                                          minsize = 20)
#                         TFs <- as.data.frame(TFs)
#                         TFs <- TFs[which(TFs$statistic == "norm_wmean"),c(2,4)]
#                         as_input <- TFs[,2]
#                         names(as_input) <- TFs[,1]
#                         return(as_input)
#                         })
```

T48:
```{r}
# RNA_scaled <- T48_transcriptomic
# 
# T48_TF_activities <- apply(RNA_scaled,2,function(x){
#                         x <- as.data.frame(x[which(!is.na(x))])
#                         TFs <- run_wmean(as.matrix(x),
#                                          network = unique(ci_regulons),
#                                          times = 1000,
#                                          minsize = 20)
#                         TFs <- as.data.frame(TFs)
#                         TFs <- TFs[which(TFs$statistic == "norm_wmean"),c(2,4)]
#                         as_input <- TFs[,2]
#                         names(as_input) <- TFs[,1]
#                         return(as_input)
#                         })
```

#### ULM
T0
```{r}
RNA_scaled <- T0_transcriptomic

T0_TF_activities <- apply(RNA_scaled,2,function(x){
                        x <- as.data.frame(x[which(!is.na(x))]) # no NAs allowed
                        TFs <- run_ulm(as.matrix(x), # actual running of TF activity estimation function
                                         network = unique(ci_regulons),
                                         # times = 1000,
                                         minsize = 20)
                        TFs <- as.data.frame(TFs) # Reformatting into cosmos input format
                        TFs <- TFs[which(TFs$statistic == "ulm"),c(2,4)]
                        as_input <- TFs[,2]
                        names(as_input) <- TFs[,1]
                        return(as_input)
                        })
```
T48
```{r}
RNA_scaled <- T48_transcriptomic

T48_TF_activities <- apply(RNA_scaled,2,function(x){
                        x <- as.data.frame(x[which(!is.na(x))]) # no NAs allowed
                        TFs <- run_ulm(as.matrix(x), # actual running of TF activity estimation function
                                         network = unique(ci_regulons),
                                         # times = 1000,
                                         minsize = 20)
                        TFs <- as.data.frame(TFs) # Reformatting into cosmos input format
                        TFs <- TFs[which(TFs$statistic == "ulm"),c(2,4)]
                        as_input <- TFs[,2]
                        names(as_input) <- TFs[,1]
                        return(as_input)
                        })
```

#### MLM
T0
```{r}
# RNA_scaled <- T0_transcriptomic
# 
# T0_TF_activities <- apply(RNA_scaled,2,function(x){
#                         x <- as.data.frame(x[which(!is.na(x))]) # no NAs allowed
#                         TFs <- run_mlm(as.matrix(x), # actual running of TF activity estimation function
#                                          network = unique(ci_regulons_no_cor),
#                                          # times = 1000,
#                                          minsize = 20)
#                         TFs <- as.data.frame(TFs) # Reformatting into cosmos input format
#                         TFs <- TFs[which(TFs$statistic == "mlm"),c(2,4)]
#                         as_input <- TFs[,2]
#                         names(as_input) <- TFs[,1]
#                         return(as_input)
#                         })
```
T48
```{r}
# RNA_scaled <- T48_transcriptomic
# 
# T48_TF_activities <- apply(RNA_scaled,2,function(x){
#                         x <- as.data.frame(x[which(!is.na(x))]) # no NAs allowed
#                         TFs <- run_mlm(as.matrix(x), # actual running of TF activity estimation function
#                                          network = unique(ci_regulons_no_cor),
#                                          # times = 1000,
#                                          minsize = 20)
#                         TFs <- as.data.frame(TFs) # Reformatting into cosmos input format
#                         TFs <- TFs[which(TFs$statistic == "mlm"),c(2,4)]
#                         as_input <- TFs[,2]
#                         names(as_input) <- TFs[,1]
#                         return(as_input)
#                         })
```


### Formatting COSMOS inputs
Reformatting gene expression data
T0
```{r}
T0_transcriptomic <- apply(T0_transcriptomic, 2, function(x){
  return(x[which(!is.na(x))])
}, simplify = F)

```
T48
```{r}
T48_transcriptomic <- apply(T48_transcriptomic, 2, function(x){
  return(x[which(!is.na(x))])
}, simplify = F)
```

Reformatting Metabolite data:
T0
```{r}
row.names(T0_metabolomic) <- T0_metabolomic$metabolite
T0_metabolomic <- T0_metabolomic[,-1]
T0_metabolomic <- apply(T0_metabolomic, 2, function(x){
  x <- x[which(!is.na(x))]
  return(x)
},simplify = F)
```
T48
```{r}
row.names(T48_metabolomic) <- T48_metabolomic$metabolite
T48_metabolomic <- T48_metabolomic[,-1]
T48_metabolomic <- apply(T48_metabolomic, 2, function(x){
  x <- x[which(!is.na(x))]
  return(x)
},simplify = F)
```

## Cosmos_inputs end product

T0:
```{r}
TF_activities <- T0_TF_activities
RNA_scaled <- T0_transcriptomic
metabs_scaled <- T0_metabolomic


T0_cosmos_inputs <- sapply(names(TF_activities), function(cell_line, RNA_scaled, metabs_scaled, TF_activities){
  cosmos_input <- list()
  cosmos_input[["RNA"]] <- RNA_scaled[[cell_line]]
  cosmos_input[["metabolomic"]] <- metabs_scaled[[cell_line]]
  cosmos_input[["TF_scores"]] <- TF_activities[[cell_line]]
  return(cosmos_input)
}, RNA_scaled = RNA_scaled, metabs_scaled = metabs_scaled, TF_activities= TF_activities, USE.NAMES = T, simplify = F)

```

T48:
```{r}
TF_activities <- T48_TF_activities
RNA_scaled <- T48_transcriptomic
metabs_scaled <- T48_metabolomic


T48_cosmos_inputs <- sapply(names(TF_activities), function(cell_line, RNA_scaled, metabs_scaled, TF_activities){
  cosmos_input <- list()
  cosmos_input[["RNA"]] <- RNA_scaled[[cell_line]]
  cosmos_input[["metabolomic"]] <- metabs_scaled[[cell_line]]
  cosmos_input[["TF_scores"]] <- TF_activities[[cell_line]]
  return(cosmos_input)
}, RNA_scaled = RNA_scaled, metabs_scaled = metabs_scaled, TF_activities= TF_activities, USE.NAMES = T, simplify = F)

```

## Saving data
AMC
```{r}
save(T0_cosmos_inputs, file = "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\CosmosR\\T0_cosmos_inputs.RData")
save(T48_cosmos_inputs, file = "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\CosmosR\\T48_cosmos_inputs.RData")

```

Laptop
```{r}
# save(T0_cosmos_inputs, file = "../ourdata/cosmos/T0_cosmos_inputs.RData")
# save(T48_cosmos_inputs, file = "../ourdata/cosmos/T48_cosmos_inputs.RData")

```

