---
title: "5 - Run COSMOS analysis"
author: "Cathy Magnée"
date: "2024-06-25"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Running COSMOS

This document describes the process of running COSMOS from COSMOS
inputs. This will happen for two datafiles in multiple steps. First, the
data, metanetwork and packages will be loaded into the environment.
Next, the options for CARNIVAL will be chosen and explained. CARNIVAL
will run using the COSMOS metanetwork and our data both forward
(signalling to metabolism) and back (metabolism to signalling). Finally,
the forward and backward outputs will be combined. This will happen
first for T0, then for T48.

This script is based of the run_cosmos.R document in the FACTOR_COSMOS
github by Saezlab. This document is preceded by the 4 - Preparing COSMOS
inputs R markdown document and procedes the 6 - Preprocess network R
markdown document.

This document is annotated to explain the steps and the choices that
were made during this project. Other options will be highlighted.

### Setting up directories, loading libraries and data

Installing packages:

```{r}
# make sure that CARNIVAL and COSMOS are installed from github:
# if(!require("remotes")) install.packages("remotes")
# remotes::install_github("saezlab/cosmosR")
```

Loading packages:

```{r}
start.time <- Sys.time()
library(cosmosR)
library(readr)
library(dplyr)
```

Data path: AMC:
```{r}
T0_input_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\CosmosR\\T0_cosmos_inputs.RData"
T48_input_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\CosmosR\\T48_cosmos_inputs.RData"
```

Laptop:
```{r}
# T0_input_path <- "../ourdata/cosmos/T0_cosmos_inputs.RData"
# T48_input_path <- "../ourdata/cosmos/T48_cosmos_inputs.RData"
```
Adjust the above chunks accordingly

Loading data:

```{r}
data("meta_network")

meta_network <- meta_network_cleanup(meta_network)

load(T0_input_path)
load(T48_input_path)

```

### Defining Cplex options
In the code block below, first the my_options list is created. If you
have a Linux OS, this first line is all you need to run. If not: Same.
Adjust the options so they fit your requirements. This is the part where
you must have the IBM Cplex Solver. It is found through academic license
here:
<https://community.ibm.com/community/user/ai-datascience/blogs/xavier-nodet1/2020/07/09/cplex-free-for-students>
It is quite a process to get the installation, as you will also need
IBMs Download Director (DD) and Java. However, the installer walks you
through this nicely.

Once you have IBM Cplex solver, adjust the location (though I think it
should be similar to mine).
```{r}

my_options <- default_CARNIVAL_options(solver = "cplex")
my_options$solverPath <- "C:/Program Files/IBM/ILOG/CPLEX_Studio2211/cplex/bin/x64_win64"
my_options$workdir <- "C:/Users/cmagnee/Documents/cplex"
my_options$outputFolder <- "C:/Users/cmagnee/Documents/cplex"
my_options$solver <- "cplex"
my_options$mipGAP <- 0.05
my_options$poolCap <- 100
my_options$limitPop <- 500
my_options$threads <- 19 # Adjust based on your machine
# my_options$threads <- 6 # My laptop
my_options$timelimit <- 3600  

```

### Defining functions:

Note for filtering input: Metabolites: Choose which compartment to
assign to the metabolic measurements: Possible values are "r", "c", "e",
"x", "m", "l", "n" and "g". More info here
<http://bigg.ucsd.edu/compartments>

"r", endoplasmic reticulum

"c", cytosol

"e", extracellular space

"x", peroxisome / glyoxysome

"m", mitochondria

"l", lysosome

"n", nucleus

"g", golgi apparatus


### Defining variables, create dfs
```{r}

max_depth <- 3
seconds_per_step <- 3600

time_df <- data.frame()

```

The chunk below runs cosmos for both timepoints, according to the variables you set in the chunks above. 
Do not forget to edit the filter accordingly!
```{r}
start.time <- Sys.time()
for (i in 1:2){
  cosmos_inputs <- list(T0_cosmos_inputs, T48_cosmos_inputs)[[i]]
  stim_point <- list("T0", "T48")[[i]]
  for (patient in names(cosmos_inputs)){
    skip_to_next <<- F
    tryCatch({
      sig_input <- cosmos_inputs[[patient]]$TF_scores
      metab_input <- cosmos_inputs[[patient]]$metabolomic 
      RNA_input <- cosmos_inputs[[patient]]$RNA
    
      # Filter inputs
      metab_input <- prepare_metab_inputs(metab_input, c("c","m"))
      
      
      filter <- "s10m05" ## This one is important!! s = signalling input, m = metabolite input filter do not put in the points or smtng
      sig_input <- sig_input[abs(sig_input) > 1.0] 
      metab_input <- metab_input[abs(metab_input) > 0.5] 
      
      metab_input <- cosmosR:::filter_input_nodes_not_in_pkn(metab_input, meta_network)
      sig_input <- cosmosR:::filter_input_nodes_not_in_pkn(sig_input, meta_network)
      
      # Stp 1
      test_for <- preprocess_COSMOS_signaling_to_metabolism(meta_network = meta_network[,c(1,3,2)],
                                                            signaling_data = sig_input,
                                                            metabolic_data = metab_input,
                                                            diff_expression_data = RNA_input,
                                                            maximum_network_depth = max_depth,
                                                            remove_unexpressed_nodes = T,
                                                            filter_tf_gene_interaction_by_optimization = T,
                                                            CARNIVAL_options = my_options)
      my_options$timelimit <- seconds_per_step
      new_row <- list(
        patient = patient,
        step = "step_1",
        end_time = round(Sys.time()-start.time,2)
      )
      time_df <- rbind(time_df, new_row)
      #Stp 2
      test_result_for <- run_COSMOS_signaling_to_metabolism(data = test_for,
                                                            CARNIVAL_options = my_options)
      new_row <- list(
        patient = patient,
        step = "step_2",
        end_time = round(Sys.time()-start.time,2)
      )
      time_df <- rbind(time_df, new_row)
      
      formatted_res <- format_COSMOS_res(test_result_for)
      
      SIF <- formatted_res[[1]]
      ATT <- formatted_res[[2]]
      
      SIF <- SIF[which(SIF$Weight != 0),]
      
      RNA_input_df <- data.frame(Nodes = names(RNA_input), t = RNA_input)
      ATT <- merge(ATT, RNA_input_df, all.x = T)
      ATT <- ATT[ATT$AvgAct != 0,]
      
      
      # write_csv(SIF, file = paste("../results/",paste(patient, "_SIF.csv",sep = ""), sep = ""))
      # write_csv(ATT, file = paste("../results/",paste(patient, "_ATT.csv",sep = ""), sep = ""))
      
      #Stp 3
      my_options$timelimit <- seconds_per_step
      
      
      test_back <- preprocess_COSMOS_metabolism_to_signaling(meta_network = meta_network[,c(1,3,2)],
                                                             signaling_data = sig_input,
                                                             metabolic_data = metab_input,
                                                             diff_expression_data = RNA_input,
                                                             maximum_network_depth = max_depth,
                                                             remove_unexpressed_nodes = TRUE,
                                                             filter_tf_gene_interaction_by_optimization = TRUE,
                                                             CARNIVAL_options = my_options)
      new_row <- list(
        patient = patient,
        step = "step_3",
        end_time = round(Sys.time()-start.time,2)
      )
      time_df <- rbind(time_df, new_row)
      #Stp 4
      my_options$timelimit <- seconds_per_step
      
      test_result_back <- run_COSMOS_metabolism_to_signaling(data = test_back,
                                                             CARNIVAL_options = my_options)
      new_row <- list(
        patient = patient,
        step = "step_4",
        end_time = round(Sys.time()-start.time,2)
      )
      time_df <- rbind(time_df, new_row)
      formatted_res_back <- format_COSMOS_res(test_result_back)
      
      SIF_back <- formatted_res_back[[1]]
      ATT_back <- formatted_res_back[[2]]
      
      SIF_back <- SIF_back[which(SIF_back$Weight != 0),]
      
      ATT_back <- merge(ATT_back, RNA_input_df, all.x = T)
      ATT_back <- ATT_back[ATT_back$AvgAct != 0,]
      
      # write_csv(SIF_back, file = paste("../results/",paste(patient, "_SIF_back.csv",sep = ""), sep = ""))
      # write_csv(ATT_back, file = paste("../results/",paste(patient, "_ATT_back.csv",sep = ""), sep = ""))
      
      #Stp 5: Merge all & save
      
      SIF_full <- as.data.frame(rbind(SIF,SIF_back))
      SIF_full <- unique(SIF_full)
      
      ATT_full <- as.data.frame(rbind(ATT,ATT_back))
      ATT_full <- unique(ATT_full)
      
      P_nodes <- ATT_full[ATT_full$NodeType == "P","Nodes"]
      M_nodes <- ATT_full[ATT_full$NodeType == "M","Nodes"]
      C_nodes <- union(P_nodes,M_nodes)
      # C_nodes <- C_nodes[which(C_nodes %in% c(SIF$Node1,SIF$Node2) & C_nodes %in% c(SIF_back$Node1,SIF_back$Node2))]
      
      ATT_full <- ATT_full[,-6]
      
      ATT_full <- ATT_full %>% group_by(Nodes) %>% summarise_each(funs(mean(., na.rm = TRUE)))
      ATT_full <- as.data.frame(ATT_full)
      
      ATT_full$NodeType <- ifelse(ATT_full$Nodes %in% C_nodes,"C",ifelse(ATT_full$Nodes %in% P_nodes,"P",ifelse(ATT_full$Nodes %in% M_nodes,"M","")))
      
      # Automatically create path to save results, so it does not get messy in my folders
      core_path <- "L:/basic/divg/EXIM/ImmunoHematology/Cathy Magnée/Data/CD3_HDvsCLL/CosmosR"
      
      result_path <- paste("/Networks/d", max_depth, sep = "") # Add depth folder
      dir.create(file.path(core_path, result_path), showWarnings = FALSE)
      result_path <- paste(result_path, paste("/f", filter, sep = ""), sep = "") # Add filter folder
      dir.create(file.path(core_path, result_path), showWarnings = FALSE)
      result_path <- paste(result_path, paste("/", stim_point, sep = ""), sep = "") # Add stim/unstim point folder
      dir.create(file.path(core_path, result_path), showWarnings = FALSE)
      result_path <- paste(result_path, "/csvs/", sep = "") # Add csv folder
      dir.create(file.path(core_path, result_path), showWarnings = FALSE)
      
      result_path <- paste(core_path, result_path, sep = "")
      # result_path <- paste("../results/Network_files_csvs/d", paste(max_depth, paste("/f", paste (filter, paste("/"), paste(stim_point, "/", sep = ""), sep = ""), sep = ""), sep = ""), sep = "")
      write_csv(SIF_full, file = paste(result_path, paste(patient, "_SIF_full.csv", sep = ""), sep = ""))
      write_csv(ATT_full, file = paste(result_path, paste(patient, "_ATT_full.csv", sep = ""), sep = ""))
      
      # How long did this take?
      new_row <- list(
        patient = patient,
        step = 'final',
        end_time = round(Sys.time()-start.time,2))
      time_df <- rbind(time_df, new_row)
    }, error = function(e) {
      message(paste("Error in processing patient:", patient, "-", e$message))
      skip_to_next <<- TRUE  # Set flag to skip the rest of the iteration
    })
    if (skip_to_next){
      next
    }
  }
}
end.time <- Sys.time()
time.taken <- round(end.time - start.time, 2)
time.taken
```



### Optional: Calculating input per filtering input

The chunk below calculates how many signalling inputs and metabolic
inputs are left after filtering for each patients.

```{r}
# ## Chunk to check how much data is available per sample
# cosmos_inputs <- T48_cosmos_inputs # T0 or T48
# 
# inputs <- data.frame()
# 
# for (patient in names(cosmos_inputs)){
#   sig_input <- cosmos_inputs[[patient]]$TF_scores
#   metab_input <- cosmos_inputs[[patient]]$metabolomic
# 
#   RNA_input <- cosmos_inputs[[patient]]$RNA
#   metab_input <- prepare_metab_inputs(metab_input, c("c","m"))
# 
#   sig_input <- sig_input[abs(sig_input) > 1.5]
#   metab_input <- metab_input[abs(metab_input) > 1.0]
# 
#   metab_input <- cosmosR:::filter_input_nodes_not_in_pkn(metab_input, meta_network)
#   sig_input <- cosmosR:::filter_input_nodes_not_in_pkn(sig_input, meta_network)
#   new_row <- list(
#     patient = patient,
#     n_sig_inpul = length(sig_input),
#     n_metab_input = length(metab_input)
#   )
#   inputs <- rbind(inputs, new_row)
# }

```
