---
title: "6-Analyze_Network"
author: "Cathy Magnée"
date: "2024-07-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyze networks

This documents is the sixth in this project. The first goal is to make the network files Cytoscape-ready. The first things to tackle are issues in uniformity and duplicates.

#### Loading

```{r}
library(readr)
library(dplyr)
library(stringr)
```

Set variables, what networks do you want to preprocess?
```{r}
# edit these based on variables just to create network
max_depth <- 3
stim_point <- "T48"
filter <- "s10m05" 
```

Setting paths based on the above set variables:
```{r}
# Path to where the data should be at AMC pc:
core_path <- "L:/basic/divg/EXIM/ImmunoHematology/Cathy Magnée/Data/CD3_HDvsCLL/CosmosR"

result_path <- paste("/Networks/d", max_depth, sep = "") # Add depth folder
result_path <- paste(result_path, paste("/f", filter, sep = ""), sep = "") # Add filter folder
result_path <- paste(result_path, paste("/", stim_point, sep = ""), sep = "") # Add stim/unstim point folder
result_path <- paste(result_path, "/csvs/", sep = "") # Add csv folder

data_path <- paste(core_path, result_path, sep = "")

# Edit below to what you want to preprocess
all_files <- list.files(path = data_path, pattern = "*_SIF_full.csv", full.names = TRUE)
```

Loop through files available at specified location. In this loop, the following things happen:
- MoleculeType column is added
- Nodenames are adjusted
- removal of duplicate nodes and edges (summarized to mean)
Things that you can uncomment if you want, but I do not recommend:
- Remove _c and _m suffixes (which stand for cytosol and mitochondrial respectively)
  Do not remove if you want to keep this info in the network
- Add Activity_ and Measured_In_ columns that are unique to the patient/sample
  Useful if you import into cytoscape directly and want to merge the networks there. Not useful if you use 7 - Cytoscape.Rmd
  
```{r, include=FALSE}
for (files in all_files){
  patient <- str_extract(files, "[^/]+(?=_SIF)")
  
  # Importing SIF and ATT file for each available patient / HD sample
  SIF_path <- paste(data_path, paste(patient, "_SIF_full.csv", sep = ""), sep = "")
  SIF <- read_csv(SIF_path)
  SIF <- unique(SIF)

  ATT_path <- paste(data_path, paste(patient, "_ATT_full.csv",sep = ""), sep = "")
  ATT <- read_csv(ATT_path)
  ATT <- unique(ATT)
  
  # Creating MoleType df for Metabs 
  MolType_df <- data.frame()
  for(nodes in ATT$Nodes){
    # print(substr(nodes, 0, 5))
    if(substr(nodes, 0, 5) == "Metab"){
      new_row <- list(Node = nodes,
                      MoleculeType = "Metabolite")
      MolType_df <- rbind(MolType_df, new_row)
    } else if(substr(nodes, 0, 5) == "Enzym"){
      new_row <- list(Node = nodes,
                      MoleculeType = "Enzyme")
      MolType_df <- rbind(MolType_df, new_row)
    } else {
      new_row <- list(Node = nodes,
                      MoleculeType = NA)
      MolType_df <- rbind(MolType_df, new_row)
    }
    
  }
  # Creating MoleculeType column. This will be built on in 7 - Cytoscape.Rmd
  ATT <- merge(ATT, MolType_df, by.x = "Nodes", by.y = "Node", all.x = T)
  
  # Sub Enzyme0000__ (random numbers) to Enzyme_
  SIF$Node1 <- sub("Enzyme[0-9]+__", "Enzyme_", SIF$Node1)
  SIF$Node2 <- sub("Enzyme[0-9]+__", "Enzyme_", SIF$Node2)
  ATT$Nodes <- sub("Enzyme[0-9]+__", "Enzyme_", ATT$Nodes)

  #Filter these rando Enzymes to collaborate (all become 1, mean of availables)
  ATT <- ATT %>%
    group_by(Nodes) %>% # Group by the Nodes column
    summarize(across(everything(), mean, na.rm = TRUE))

  # Remove Metab__ part of Node names
  SIF$Node1 <- sub(".*__", "", SIF$Node1)
  SIF$Node2 <- sub(".*__", "", SIF$Node2)
  ATT$Nodes<- sub(".*__", "", ATT$Nodes)
  
  SIF$Node1 <- sub(',', "", SIF$Node1)
  SIF$Node2 <- sub(',', "", SIF$Node2)
  ATT$Nodes<- sub(',', "", ATT$Nodes)
  SIF$Node1 <- sub(',', "", SIF$Node1)
  SIF$Node2 <- sub(',', "", SIF$Node2)
  ATT$Nodes<- sub(',', "", ATT$Nodes)
  
  SIF <- SIF |>
    filter(Node1 != Node2) |> # First removing nodes that are re-activating themselves
    group_by(Node1, Sign, Node2) |>
    summarize(Weight = sum(Weight), .groups='drop')
  
  # Remove _c and _m from Node names (metabolites). But write down in NodeType # Cancelled because it gives information about cell system.
  # ATT$NodeType <- ifelse(grepl("_c$", ATT$Nodes) & is.na(ATT$NodeType), "C", ATT$NodeType)
  # ATT$NodeType <- ifelse(grepl("_m$", ATT$Nodes), "M", ATT$NodeType)
  # 
  # SIF$Node1 <- sub("_m", "", SIF$Node1)
  # SIF$Node2 <- sub("_m", "", SIF$Node2)
  # ATT$Nodes<- sub("_m", "", ATT$Nodes)
  # 
  # SIF$Node1 <- sub("_c", "", SIF$Node1)
  # SIF$Node2 <- sub("_c", "", SIF$Node2)
  # ATT$Nodes<- sub("_c", "", ATT$Nodes)
  
  # # Add additional column "Acticity_ (patient network name)". This will make network processing within cytoscape easier.
  # temp <- paste("Activity_", patient, "")
  # ATT[temp] <- ATT$Activity
  # 
  # # Add additional column "In_(patient network name)". This will make network processing within cytoscape easier.
  # temp <- paste("In_", patient,sep = "")
  # ATT[temp] <- 1
 
  # Saving processed csvs
  # result_path <- paste("/processed_results/d", max_depth, sep = "") # Laptop only
  data_path <- paste(core_path, result_path, sep = "") # Laptop only
  write_csv(SIF, file = paste(data_path, paste(patient, "_SIF_processed.csv", sep = ""), sep = ""))
  write_csv(ATT, file = paste(data_path, paste(patient, "_ATT_processed.csv", sep = ""), sep = ""))
  
}

```

