---
title: "6-Analyze_Network"
author: "Cathy Magnée"
date: "2024-07-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyze networks

This documents is the sixth in this project. The first goal is to make the network files Cytoscape-ready. The first things to tackle are issues in uniformity and duplicates.

#### Loading

```{r}
library(readr)
library(dplyr)
```

I am not sure yet to which extent I want to edit these networks one-by-one, or automatically. For now, we do one-by-one. Probably will automate later x

Input which network you want to edit
```{r}
patient <- "CLL2479_T48" # Pick a person for who you want to make the network
max_depth <- 2
```

Input paths AMC and laptop are the same:
```{r}
SIF_path <- paste("../results/d", paste(max_depth, paste(patient, "_SIF_full.csv", sep = ""), sep = ""), sep = "")
ATT_path <- paste("../results/d", paste(max_depth, paste(patient, "_ATT_full.csv",sep = ""), sep = ""), sep = "")
```
But not all! AMC:
```{r}
# AMC
# T0_input_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\CosmosR\\T0_cosmos_inputs.RData"
# T48_input_path <- "L:\\basic\\divg\\EXIM\\ImmunoHematology\\Cathy Magnée\\Data\\CD3_HDvsCLL\\CosmosR\\T48_cosmos_inputs.RData"
```
Laptop
```{r}
T0_input_path <- "../ourdata/cosmos/T0_cosmos_inputs.RData"
T48_input_path <- "../ourdata/cosmos/T48_cosmos_inputs.RData"
```


```{r}
SIF <- read_csv(SIF_path)
ATT <- read_csv(ATT_path)
```
### Uniformity
Removing the everything and the "__" from the names in columns $Node1 and $Node2

```{r}
SIF$Node1 <- sub(".*__", "", SIF$Node1)
SIF$Node2 <- sub(".*__", "", SIF$Node2)
ATT$Nodes<- sub(".*__", "", ATT$Nodes)

SIF$Node1 <- sub("_c", "", SIF$Node1)
SIF$Node2 <- sub("_c", "", SIF$Node2)
ATT$Nodes<- sub("_c", "", ATT$Nodes)

SIF$Node1 <- sub("_m", "", SIF$Node1)
SIF$Node2 <- sub("_m", "", SIF$Node2)
ATT$Nodes<- sub("_m", "", ATT$Nodes)


```

Remove duplicates and sum up their weights
```{r}
SIF <- SIF |>
  filter(Node1 != Node2) |> # First removing nodes that are re-activating themselves
  group_by(Node1, Sign, Node2) |>
  summarize(Weight = sum(Weight), .groups='drop')

```

```{r}
write_csv(SIF, file = paste("../results/processed_results/d", paste(max_depth, paste(patient, "_SIF_full.csv", sep = ""), sep = ""), sep = ""))
write_csv(ATT, file = paste("../results/processed_results/d", paste(max_depth, paste(patient, "_ATT_full.csv",sep = ""), sep = ""), sep = ""))
```



Now in a loop:
```{r, include=FALSE}

load(T0_input_path)
load(T48_input_path)
T48_cosmos_inputs <- day2_cosmos_inputs

# Edit below to what you want to preprocess
cosmos_inputs <- T48_cosmos_inputs # edit these as please you
max_depth <- 2 # edit these as please you


for (patient in names(cosmos_inputs)){
  # insert paths
  SIF_path <- paste("../results/d", paste(max_depth, paste(patient, "_SIF_full.csv", sep = ""), sep = ""), sep = "")
  SIF <- read_csv(SIF_path)

  ATT_path <- paste("../results/d", paste(max_depth, paste(patient, "_ATT_full.csv",sep = ""), sep = ""), sep = "")
  ATT <- read_csv(ATT_path)
  
  SIF$Node1 <- sub(".*__", "", SIF$Node1)
  SIF$Node2 <- sub(".*__", "", SIF$Node2)
  ATT$Nodes<- sub(".*__", "", ATT$Nodes)
  
  SIF$Node1 <- sub(',', "", SIF$Node1)
  SIF$Node2 <- sub(',', "", SIF$Node2)
  ATT$Nodes<- sub(',', "", ATT$Nodes)
  SIF$Node1 <- sub(',', "", SIF$Node1)
  SIF$Node2 <- sub(',', "", SIF$Node2)
  ATT$Nodes<- sub(',', "", ATT$Nodes)
  
  # SIF$Node1 <- sub("_m", "", SIF$Node1)
  # SIF$Node2 <- sub("_m", "", SIF$Node2)
  # ATT$Nodes<- sub("_m", "", ATT$Nodes)
  
  SIF <- SIF |>
  filter(Node1 != Node2) |> # First removing nodes that are re-activating themselves
  group_by(Node1, Sign, Node2) |>
  summarize(Weight = sum(Weight), .groups='drop')
  
  write_csv(SIF, file = paste("../results/processed_results/d", paste(max_depth, paste(patient, "_SIF_full.csv", sep = ""), sep = ""), sep = ""))
  write_csv(ATT, file = paste("../results/processed_results/d", paste(max_depth, paste(patient, "_ATT_full.csv",sep = ""), sep = ""), sep = ""))
  
}
# networks.path <- "../results/"
# all_files <- list.files(path = networks.path, pattern = "*._full.csv", full.names = TRUE)


```

