---
title: "6-Analyze_Network"
author: "Cathy Magnée"
date: "2024-07-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyze networks

This documents is the sixth in this project. The first goal is to make the network files Cytoscape-ready. The first things to tackle are issues in uniformity and duplicates.

#### Loading

```{r}
library(readr)

```

I am not sure yet to which extent I want to edit these networks one-by-one, or automatically. For now, we do one-by-one. Probably will automate later x

Input which network you want to edit
```{r}
patient <- "CLL2479_T48" # Pick a person for who you want to make the network
max_depth <- 2
```



Input paths AMC and laptop are the same:
```{r}
SIF_path <- paste("../results/d", paste(max_depth, paste(patient, "_SIF_full.csv", sep = ""), sep = ""), sep = "")
ATT_path <- paste("../results/d", paste(max_depth, paste(patient, "_ATT_full.csv",sep = ""), sep = ""), sep = "")
```

```{r}
SIF <- read_csv(SIF_path)
ATT <- read_csv(ATT_path)
```
### Uniformity
Removing the everything and the "__" from the names in columns $Node1 and $Node2

```{r}
SIF$Node1 <- sub(".*__", "", SIF$Node1)
SIF$Node2 <- sub(".*__", "", SIF$Node2)
ATT$Nodes<- sub(".*__", "", ATT$Nodes)

```

Remove duplicates and sum up their weights
```{r}
SIF <- SIF |>
  filter(Node1 != Node2) |> # First removing nodes that are re-activating themselves
  group_by(Node1, Sign, Node2) |>
  summarize(Weight = sum(Weight), .groups='drop')

```


```{r}
write_csv(SIF, file = paste("../results/d", paste(max_depth, paste(patient, "_SIF_full.csv", sep = ""), sep = ""), sep = ""))
write_csv(ATT, file = paste("../results/d", paste(max_depth, paste(patient, "_ATT_full.csv",sep = ""), sep = ""), sep = ""))
```




