---
title: "7 - Cytoscape"
author: "Cathy Magnée"
date: "2024-09-10"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Cytoscape

This R Markdown document is the seventh in the EXIM COSMOS project
pipeline, where multi-omics data is transformed into networks to compare
between the T cell mechanistic networks of healthy donors and CLL
patients. The goal of this document is to transform, merge and visualize
the networks created in step 5 and processed in step 6 of the pipeline.

### Loading libraries

```{r}
library(RCy3)
library(readr)
library(stringr)
library(dplyr)
```

## Creating functions

Three functions are utilized in this script. The first one processes and
merges the available files, second one filtered the merged files, the third one creates the network.

Function 1: Process and merge files:
This function loops through the available network files (based on variables that you can edit below), checks if it is a CLL or HD network, then merges accordingly. It creates column Measured, which counts in how many networks this node is present. 
```{r}
# Define the process_files function with a single customizable filter function
merge_files <- function(all_files, data_path, group_type) {
  Merged_SIF <- data.frame()
  Merged_ATT <- data.frame()
  for (files in all_files){
    # Extract patient information from file name
    patient <- str_extract(files, "[^/]+(?=_SIF)")
    
    # Import processed SIF and ATT files for the patient
    SIF_path <- paste(data_path, paste(patient, "_SIF_processed.csv", sep = ""), sep = "")
    SIF <- read_csv(SIF_path, show_col_types = FALSE) |> unique()
    
    ATT_path <- paste(data_path, paste(patient, "_ATT_processed.csv", sep = ""), sep = "")
    ATT <- read_csv(ATT_path, show_col_types = FALSE) |> unique()
    
    # Check if CLL or HD patient
    if (startsWith(patient, group_type)){
      # Edges / SIF part
      for (i in 1:nrow(SIF)) { 
        Edge <- SIF[i, ] 
        Merged_SIF <- rbind(Merged_SIF, Edge) 
      } # Merge result: All edges for all CLL patiens
      
      # Nodes / ATT part
      for (Noderow in unique(ATT$Nodes)) {
        # If node already in merged:
        if (Noderow %in% Merged_ATT$Nodes) { 
          # Show in Measured that it is Measured in +1 sample
          Merged_ATT$Measured[Merged_ATT$Nodes == Noderow] <- 
            Merged_ATT$Measured[Merged_ATT$Nodes == Noderow] + 1 
          # Also activity == mean activity between what is available
          Merged_ATT$Activity[Merged_ATT$Nodes == Noderow] <- (Merged_ATT$Activity[Merged_ATT$Nodes == Noderow] + mean(
            ATT$Activity[ATT$Nodes == Noderow])) #/ Merged_ATT$Measured[Merged_ATT$Nodes == Noderow] 
        } else { # If node does not exists in Merged yet (new node)
          # Add node to merged df
          new_row <- list(
            Nodes = Noderow,
            Measured = 1,
            measured = mean(ATT$measured[ATT$Nodes == Noderow]),
            MoleculeType = ATT$MoleculeType[ATT$Nodes == Noderow][1],
            Activity = mean(ATT$Activity[ATT$Nodes == Noderow])
          )
          Merged_ATT <- rbind(Merged_ATT, new_row)
        }
      }
    }
  }
  Merged_ATT$Activity <- Merged_ATT$Activity/Merged_ATT$Measured
  return(list(
    Merged_ATT = Merged_ATT,
    Merged_SIF = Merged_SIF
  ))
}


```

Function 2: Filter functions

Note: Also includes filter functions. Filter functions encompass the
decisions of what nodes you want to keep when merging the network. The
decision is between:

-   max:

    Only the nodes present in all sample networks will be kept (very
    strict).

-   minus:

    Nodes present in all but one of the available networks will be kept.

-   majority:

    All nodes present in the majority of the network (more than half)
    will be kept (least strict).

-   min:

    Nodes present in \>= the minimum amount of samples available will be
    kept.

Now, the code:
```{r}

filter_merged_files <- function(ATT, SIF, filter_function){
  Merged_ATT <- unique(ATT)
  Merged_SIF <- unique(SIF)

  Filtered_Merged_ATT <- filter_function(Merged_ATT)
  
  # Only keep edges for which both nodes are present in the network
  Merged_SIF <- Merged_SIF |>
    filter(Node1 %in% Filtered_Merged_ATT$Nodes) |>
    filter(Node2 %in% Filtered_Merged_ATT$Nodes)
  
   # Summarize edge weights
  Merged_SIF <- Merged_SIF |>
    group_by(Node1, Node2, Sign) |>
    summarize(Weight = mean(Weight), .groups = 'drop')
  
  return(list(
    Filtered_Merged_ATT = Filtered_Merged_ATT,
    Merged_SIF = Merged_SIF))
}
# Define filter functions
minus_filter_function <- function(Merged_ATT) {
  filter(Merged_ATT, Measured >= max(Measured)-1)
}

# Majority filter
majority_filter_function <- function(Merged_ATT) {
  filter(Merged_ATT, Measured > max(Measured)/ 2)
}

# Define the max filter function (Measured == max(Measured))
max_filter_function <- function(Merged_ATT) {
  filter(Merged_ATT, Measured == max(Measured))
}

# Extra filter: Keeps nodes
min_filter_function <- function(Merged_ATT) {
  filter(Merged_ATT, Measured >= min_samples)
}

# To call the function with majority filter:
# result <- process_files(all_files, data_path, majority_filter_function)

# To call the  function with minus filter:
# result <- process_files(all_files, data_path, minus_filter_function)

# To call the function with max filter:
# result <- process_files(all_files, data_path, max_filter_function)


```

Function 3: Create networks in cytoscape:

```{r}
Create_Network <- function(ATT, SIF, MergedGroup){
  # Translate this to Node Table and Edges Table for in Cytoscape
  nodes <- data.frame(id = ATT$Nodes,
                      Activity = ATT$Activity,
                      measured = ATT$measured,
                      moleculeType = ATT$MoleculeType,
                      stringsAsFactors = F)
  edges <- data.frame(source = SIF$Node1,
                      target = SIF$Node2,
                      interaction = as.character(SIF$Sign),
                      weight = SIF$Weight,
                      stringsAsFactors = FALSE)
  
  # Two things: 1: Adding moleculeType column to nodes table; 2: putting shapes to these
  # Step 1 should be done earlier, maybe in 6 - Preprocess_network.Rmd.
  #StringDB Tryout
  ATT_string_interaction_cmd <- paste('string protein query taxonID=9606 cutoff=0.9 query=', paste(nodes$id, collapse=","),'"',sep="")
  commandsGET(ATT_string_interaction_cmd )
  string_nodes <- getTableColumns(table = "node", columns = c("display name", "target::family"), network = "STRING network")
  colnames(string_nodes) <- c("id", "moleculeType")
  write.csv(string_nodes, file = paste(data_path, paste(MergedGroup, ".csv", sep = ""), sep = "StringDB_"))
  string_nodes <- merge(nodes, string_nodes, by = "id", all.x = T)
  
  nodes$moleculeType <- ifelse(is.na(string_nodes$moleculeType.x), #If x is NA, put in y, if x is not na, put in x. 
                                     string_nodes$moleculeType.y,  # moleculeType <- , else (in which case, x is not NA)
                                     string_nodes$moleculeType.x) # moleculeType <- x
  nodes$moleculeType[is.na(nodes$moleculeType)] <- "Other"
  print(nodes)
  
  
  closeSession(save.before.closing = F)

  
  createNetworkFromDataFrames(nodes, edges, title = MergedGroup, collection = MergedGroup) # if(substr(MergedGroup, 0, 3)== "CLL"){"CLL"} else { "HD"})
  # You should see a network rn frfr
  
  # Setting style for network OR creating it first, then setting the style. If else so we do not remake this style over and over.
  if (style.name == "COSMOS_Style"){
    setVisualStyle(style.name)
  } else {
    style.name = "COSMOS_Style"
    defaults <- list(NODE_SHAPE="ROUND_RECTANGLE",
                     NODE_SIZE=50,
                     EDGE_TRANSPARENCY=255)
    nodeLabels <- mapVisualProperty('node label','id','p')
    nodeFills <- mapVisualProperty('node fill color',
                                   'Activity',
                                   'c',
                                   c(-1, 1),
                                   c('#CC0000', '#009900'))
    nodeBorderWidth <- mapVisualProperty('node border width',
                                         'measured',
                                         'd',
                                         c(0.0, 1.0),
                                         c(0, 4))
    arrowShapes <- mapVisualProperty('Edge Target Arrow Shape', # What style property we are mapping
                                     'interaction', # What table column we are using for it
                                     'd', # Type of mapping (discrete)
                                     c("-1", "1"),c("T","Arrow"))
    edgeWidth <- mapVisualProperty('edge width','weight','p')
  
  
    createVisualStyle(style.name, defaults, list(nodeLabels,nodeFills,nodeBorderWidth,arrowShapes,edgeWidth))
    setVisualStyle(style.name)
  }
  
  
  # setLayoutProperties("degree-circle")
  # 2: Connecting shapes to moleculeType
  
  column <- 'moleculeType'
  values <- c ('Metabolite', 'Kinase',
               'Transcription Factor', 'Other',
               'Enzyme', 'Epigenetic',
               'GPCR', 'TF-Epigenetic',
               'Nuclear Receptor', 'Transporter')
  shapes <- c ('ELLIPSE', 'TRIANGLE',
               'DIAMOND', 'PARALLELOGRAM',
               'RECTANGLE', 'VEE',
               'HEXAGON', 'DIAMOND',
               'OCTAGON', 'ROUND RECTANGLE')
  setNodeShapeMapping (column, values, shapes, style.name = style.name)
  layoutNetwork("degree-circle")
  
  # Save session & a pdf picture of it
  saveSession(filename = paste(data_path, MergedGroup, sep = "Merged_Network_"))
  
  exportImage(paste(data_path, MergedGroup, sep = "Degree_Circle_"), 'PDF') #.pdf
  
  
  # Select nodes that have a high interconnectivity (>10 edges to other nodes)
  createDegreeFilter("ImportantNodes", criterion = c(10,100), network = MergedGroup, apply = T)
  Important_Nodes <- data.frame(getSelectedNodes())
  
  # IF there are highly interconnected nodes, safe them and their activity.
  if(length(Important_Nodes)>1){
    Important_df <- merge(Important_Nodes, Merged_ATT_CLL[c(1,5)], by.x = "getSelectedNodes..", by.y = "Nodes", all.x = T)
    names(Important_df) <- c("Nodes", "Act_CLL")
    Important_df <- merge(Important_df, Merged_ATT_HD[c(1,5)], by.x = "Nodes", by.y = "Nodes", all.x = T)
    names(Important_df) <- c("Nodes", "Act_CLL", "Act_HD")
    print(Important_df)
    
    writeLines(Important_Nodes$getSelectedNodes.., con = paste(data_path, paste(MergedGroup, ".txt", sep = ""), sep = "Important_Nodes_"))
    write.csv(Important_df, file = paste(data_path, paste(MergedGroup, ".csv", sep = ""), sep = "Important_Nodes_Act_"))
  }
}
```

### Setup Cytoscape

Make sure to have the Cytoscape software open on your device before
continuing.

```{r}
cytoscapePing()
cytoscapeVersionInfo()
style.name = "empty"
```

If the above gives an error, either wait for one minute or restart R
and/or Cytoscape.

### Setup: network Variables

Below are the variables of the networks you want to visualize.

```{r}
max_depth <- 3
filter <- "s10m05"
stim_point <- "T0"
```

### Setup: network paths

Below is the data path of my AMC pc.
```{r}
core_path <- "L:/basic/divg/EXIM/ImmunoHematology/Cathy Magnée/Data/CD3_HDvsCLL/CosmosR"
result_path <- paste("/Networks/d", max_depth, sep = "") # Add depth folder
result_path <- paste(result_path, paste("/f", filter, sep = ""), sep = "") # Add filter folder
result_path <- paste(result_path, paste("/", stim_point, sep = ""), sep = "") # Add stim/unstim point folder
result_path <- paste(result_path, "/csvs/", sep = "") # Add csv folder
data_path <- paste(core_path, result_path, sep = "")
```

And my laptop:

```{r}
# core_path <- "C:/Users/Cathy.LAPTOP-SDFOSKVI/Documents/School_files/SysBio_RP/COSMOS_EXIM/results/processed_results"
# result_path <- paste("/d", max_depth, sep = "") # Add depth folder
# data_path <- paste(core_path, result_path, sep = "")
```

Adjust this accordingly to your device and placement of files.

### Setup: other variables

```{r}
# Style name will later be replaced in creating network function
style.name = "empty"

# Place where all the files are that you will be merging then vizualizing 
all_files <- list.files(path = data_path, pattern = "*_SIF_processed.csv", full.names = TRUE)

min_samples <- min(length(which(grepl("csvs/CLL", all_files))), length(which(grepl("csvs/HD", all_files))))


```

## Merging networks:

Processing and merging networks: Choose one Default for me right now is
minus. This seems to be the most balanced option. I will also be
exploring the max function.

```{r}
# CLL Merging
result <- merge_files(all_files, data_path, "CLL")
Merged_ATT_CLL <- result$Merged_ATT
Merged_SIF_CLL <- result$Merged_SIF 

# HD Merging
result <- merge_files(all_files, data_path, "HD")
Merged_ATT_HD <- result$Merged_ATT
Merged_SIF_HD <- result$Merged_SIF 

# CLL
result <- filter_merged_files(Merged_ATT_CLL, Merged_SIF_CLL, majority_filter_function)
# result <- filter_merged_files(Merged_ATT_CLL, Merged_SIF_CLL, minus_filter_function)
# result <- filter_merged_files(Merged_ATT_CLL, Merged_SIF_CLL, max_filter_function)
# result <- filter_merged_files(Merged_ATT_CLL, Merged_SIF_CLL, min_filter_function)

Filtered_Merged_ATT_CLL <- result$Filtered_Merged_ATT
Merged_SIF_CLL <- result$Merged_SIF 

# HD
result <- filter_merged_files(Merged_ATT_HD, Merged_SIF_HD, majority_filter_function)
# result <- filter_merged_files(Merged_ATT_HD, Merged_SIF_HD, minus_filter_function)
# result <- filter_merged_files(Merged_ATT_HD, Merged_SIF_HD, max_filter_function)
# result <- filter_merged_files(Merged_ATT_HD, Merged_SIF_HD, min_filter_function)

Filtered_Merged_ATT_HD <- result$Filtered_Merged_ATT
Merged_SIF_HD <- result$Merged_SIF 


```


### Creating networks for merged:

```{r}
# CLL
Create_Network(ATT = Filtered_Merged_ATT_CLL, SIF= Merged_SIF_CLL, MergedGroup = paste(stim_point, "CLL", sep = "_"))

# HD
Create_Network(ATT = Filtered_Merged_ATT_HD, SIF= Merged_SIF_HD, MergedGroup = paste(stim_point, "HD", sep = "_"))

```

#### Creating networks for individuals:

```{r}
# patient = "CLL2105_T48" # Fill it in
# SIF_path <- paste(data_path, paste(patient, "_SIF_processed.csv", sep = ""), sep = "")
# SIF <- read_csv(SIF_path, show_col_types = F)
# ATT_path <- paste(data_path, paste(patient, "_ATT_processed.csv",sep = ""), sep = "")
# ATT <- read_csv(ATT_path, show_col_types = F)
# 
# Create_Network(ATT = ATT, SIF= SIF, MergedGroup = patient)
```
